---
title: "Persistence Analyses"
author: "McCaide Wooten"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
---
# Load Data
```{r Load Data, cache=FALSE}

library(readxl)
library(tinytex)

dat <- read_excel("C:/INSERT FILE PATHWAY HERE/Environmental Persistence Data.xlsx", sheet = "Samples")

```


# Survival Analysis
## Setup & Basic Stats
```{r Survival Analysis:Setup & Basic Stats, cache=FALSE}

library(tidyverse)
library(janitor)
library(dplyr)
library(survival)
library(stringr)
library(ggplot2)
library(Icens)
library(tibble)

# How many carcass sites have a known date/year of death? #

known_death <- dat %>%
  filter(site_id != "WOLF-23-01") %>%
  filter(carcass_id != "WOLF-23-01") %>%
  filter(carcass_association == "Y", !is.na(year_of_death))

known_death_summary <- known_death %>%
  summarise(
    n_rows = n(),
    n_carcass_id = n_distinct(carcass_id),
    n_site_id = n_distinct(site_id)
  )

known_death_summary

# Distribution of death year (unique carcass_id only) #
known_death %>%
  distinct(carcass_id, year_of_death) %>%
  count(year_of_death, sort = TRUE)

# Among known-death sites, how many have paired soil and tissue over the entire study? What is the censoring-type distribution for these soil and tissue samples? #

kd_st <- known_death %>%
  filter(substrate %in% c("Soil", "Tissue"),
         !is.na(culture_ac_pos),
         !is.na(time_since_death_years)) %>%
  mutate(
    pos = (culture_ac_pos == "Y")
  )

paired_status <- kd_st %>%
  distinct(carcass_id, substrate) %>%
  count(carcass_id, name = "n_substrates") %>%
  mutate(paired_soil_tissue = (n_substrates == 2))

paired_status %>%
  summarise(
    n_known_death_carcass = n(),
    n_paired = sum(paired_soil_tissue),
    n_unpaired = sum(!paired_soil_tissue)
  )

# List unpaired carcass_ids + which substrate they have
unpaired_list <- kd_st %>%
  distinct(carcass_id, substrate) %>%
  group_by(carcass_id) %>%
  summarise(substrates_present = paste(sort(unique(substrate)), collapse = ", "),
            .groups = "drop") %>%
  left_join(paired_status, by = "carcass_id") %>%
  filter(!paired_soil_tissue)

unpaired_list

# Censoring type per carcass_id × substrate
# Definition:
# left  = never culture positive
# right = culture positive at most recent observation
# interval = positive at some point, then negative later

# collapse to one "visit" per carcass_id × substrate × time_since_death_years
visits <- kd_st %>%
  group_by(carcass_id, substrate, time_since_death_years) %>%
  summarise(
    pos = any(pos, na.rm = TRUE),
    sample_year = min(sample_year, na.rm = TRUE),
    .groups = "drop"
  )

censor_types <- visits %>%
  arrange(carcass_id, substrate, time_since_death_years) %>%
  group_by(carcass_id, substrate) %>%
  summarise(
    any_pos = any(pos),
    last_status = dplyr::last(pos),
    censor_type = case_when(
      !any_pos ~ "left",
      last_status ~ "right",
      TRUE ~ "interval"
    ),
    .groups = "drop"
  )

# Distribution overall
censor_types %>%
  count(censor_type) %>%
  mutate(pct = round(100 * n / sum(n), 1)) %>%
  arrange(match(censor_type, c("left", "interval", "right")))

# Distribution by substrate
censor_types %>%
  count(substrate, censor_type) %>%
  group_by(substrate) %>%
  mutate(pct = round(100 * n / sum(n), 1)) %>%
  ungroup() %>%
  arrange(substrate, match(censor_type, c("left", "interval", "right")))

# Wide Table
censor_types %>%
  count(substrate, censor_type) %>%
  tidyr::pivot_wider(names_from = censor_type, values_from = n, values_fill = 0)

# Censoring distribution ONLY among paired carcasses
paired_ids <- paired_status %>%
  filter(paired_soil_tissue) %>%
  pull(carcass_id)

censor_types %>%
  filter(carcass_id %in% paired_ids) %>%
  count(substrate, censor_type) %>%
  group_by(substrate) %>%
  mutate(pct = round(100 * n / sum(n), 1)) %>%
  ungroup() %>%
  arrange(substrate, match(censor_type, c("left", "interval", "right")))

censor_types %>%
  filter(carcass_id %in% paired_ids) %>%
  count(substrate, censor_type) %>%
  tidyr::pivot_wider(names_from = censor_type, values_from = n, values_fill = 0)

# Identify left-censored carcass×substrate series (paired only) #
left_series <- censor_types %>%
  filter(carcass_id %in% paired_ids, censor_type == "left") %>%
  dplyr::select(carcass_id, substrate)

# Pull only visits that belong to left-censored series
left_visits <- visits %>%
  semi_join(left_series, by = c("carcass_id", "substrate"))

# Visit-level summary: distribution of observed sampling times (pooled across sites) #
summ_stats <- function(x) {
  tibble(
    n = sum(!is.na(x)),
    min = min(x, na.rm = TRUE),
    q1 = as.numeric(quantile(x, 0.25, na.rm = TRUE)),
    median = median(x, na.rm = TRUE),
    mean = mean(x, na.rm = TRUE),
    q3 = as.numeric(quantile(x, 0.75, na.rm = TRUE)),
    max = max(x, na.rm = TRUE)
  )
}

left_visit_summary_by_substrate <- left_visits %>%
  group_by(substrate) %>%
  summarise(summ_stats(time_since_death_years), .groups = "drop")

left_visit_summary_by_substrate

# Carcass-series summary: per carcass_id×substrate, what was the sampling window? #
left_series_windows <- left_visits %>%
  group_by(carcass_id, substrate) %>%
  summarise(
    n_visits = n(),
    first_tsd = min(time_since_death_years, na.rm = TRUE),
    last_tsd  = max(time_since_death_years, na.rm = TRUE),
    .groups = "drop"
  )

# Summarize those windows across carcasses, by substrate
left_window_summary_by_substrate <- left_series_windows %>%
  group_by(substrate) %>%
  summarise(
    n_series = n(),
    median_first = median(first_tsd, na.rm = TRUE),
    median_last  = median(last_tsd,  na.rm = TRUE),
    min_first = min(first_tsd, na.rm = TRUE),
    max_last  = max(last_tsd,  na.rm = TRUE),
    .groups = "drop"
  )

left_window_summary_by_substrate

left_series_windows %>%
  arrange(substrate, last_tsd)

##############################
# Generation of Table 1 Data #
##############################

tab_dat <- dat %>%
  filter(
    coalesce(as.character(site_id), "") != "WOLF-23-01",
    coalesce(as.character(carcass_id), "") != "WOLF-23-01"
  )

sample_types <- c("Tissue", "Soil", "Water", "Sediment", "Vegetation")

tab_dat <- dat %>%
  mutate(
    unique_id = as.character(unique_id),
    site_id = as.character(site_id),
    carcass_id = as.character(carcass_id),
    species = as.character(species),
    carcass_association = as.character(carcass_association),
    substrate = as.character(substrate),
    sample_type = substrate
  ) %>%
  filter(
    sample_type %in% sample_types,
    
    # Exclude all WOLF-associated rows robustly
    !str_detect(coalesce(unique_id, ""), "^WOLF-23-01"),
    !str_detect(coalesce(site_id, ""), "^WOLF-23-01"),
    !str_detect(coalesce(carcass_id, ""), "^WOLF-23-01"),
    
    # Remove non-carcass-associated soil samples
    !(sample_type == "Soil" & coalesce(carcass_association, "") != "Y")
  )

# Define Unique Unit for each Row #

tab_dat <- tab_dat %>%
  mutate(
    unit_id = coalesce(as.character(site_id), as.character(carcass_id))
  )

# Define Tested / Positive Flags #
tab_dat <- tab_dat %>%
  mutate(
    culture_tested = culture_done == "Y" | culture_ac_pos %in% c("Y", "N"),
    pcr_tested     = pcr_done == "Y"     | pcr_ac_pos %in% c("Y", "N"),

    culture_pos = culture_ac_pos == "Y",
    pcr_pos     = pcr_ac_pos == "Y",

    any_tested = culture_tested | pcr_tested,
    any_pos    = case_when(
      ac_any_pos %in% c("Y", "N") ~ ac_any_pos == "Y",
      TRUE ~ culture_pos | pcr_pos
    ),
      time_known = carcass_association == "Y" & !is.na(year_of_death)
  )

# Helper for n/Tested (%) Formatting #
fmt_n_tested <- function(pos, tested, digits = 1) {
  ifelse(
    tested > 0,
    paste0(pos, "/", tested, " (", round(100 * pos / tested, digits), ")"),
    ""
  )
}

fmt_n_pct <- function(n, denom, digits = 1) {
  ifelse(
    denom > 0,
    paste0(n, " (", round(100 * n / denom, digits), ")"),
    ""
  )
}

# Yearly Summaries #
by_year <- tab_dat %>%
  group_by(sample_type, sample_year) %>%
  summarise(
    samples_n      = n(),
    unique_units_n = n_distinct(unit_id),

    time_known_n = sum(time_known, na.rm = TRUE),

    culture_pos_n    = sum(culture_pos, na.rm = TRUE),
    culture_tested_n = sum(culture_tested, na.rm = TRUE),

    pcr_pos_n    = sum(pcr_pos, na.rm = TRUE),
    pcr_tested_n = sum(pcr_tested, na.rm = TRUE),

    any_pos_n    = sum(any_pos, na.rm = TRUE),
    any_tested_n = sum(any_tested, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(sample_year = as.character(sample_year))

# Add Empty Rows #
table_shell <- bind_rows(
  tibble(sample_type = "Tissue",     sample_year = c("2021", "2022", "2023", "2024")),
  tibble(sample_type = "Soil",       sample_year = c("2021", "2022", "2023", "2024")),
  tibble(sample_type = "Water",      sample_year = c("2023", "2024")),
  tibble(sample_type = "Sediment",   sample_year = c("2023", "2024")),
  tibble(sample_type = "Vegetation", sample_year = "2024")
)

table1_years <- table_shell %>%
  left_join(by_year, by = c("sample_type", "sample_year")) %>%
  mutate(
    across(
      c(samples_n, unique_units_n, time_known_n,
        culture_pos_n, culture_tested_n,
        pcr_pos_n, pcr_tested_n,
        any_pos_n, any_tested_n),
      ~ replace_na(., 0)
    )
  )

# Totals by Sample Type #
table1_totals <- tab_dat %>%
  group_by(sample_type) %>%
  summarise(
    sample_year = "TOTAL",
    samples_n      = n(),
    unique_units_n = n_distinct(unit_id),

    time_known_n = sum(time_known, na.rm = TRUE),

    culture_pos_n    = sum(culture_pos, na.rm = TRUE),
    culture_tested_n = sum(culture_tested, na.rm = TRUE),

    pcr_pos_n    = sum(pcr_pos, na.rm = TRUE),
    pcr_tested_n = sum(pcr_tested, na.rm = TRUE),

    any_pos_n    = sum(any_pos, na.rm = TRUE),
    any_tested_n = sum(any_tested, na.rm = TRUE),
    .groups = "drop"
  )

table1_out <- bind_rows(table1_years, table1_totals) %>%
  mutate(
    sample_year = as.character(sample_year)
  ) %>%
  arrange(
  factor(sample_type, levels = c("Tissue", "Soil", "Water", "Sediment", "Vegetation")),
  factor(sample_year, levels = c("2021", "2022", "2023", "2024", "TOTAL"))
) %>%
  mutate(
    Year = sample_year,

    `Samples, n` = samples_n,
    `Unique Units, n` = unique_units_n,
    `Time-Known Subset, n (%)` = fmt_n_pct(time_known_n, samples_n),

    `Culture +, +/Tested (%)` = fmt_n_tested(culture_pos_n, culture_tested_n),
    `PCR +, +/Tested (%)`     = fmt_n_tested(pcr_pos_n, pcr_tested_n),
    `Any Test +, +/Tested (%)` = fmt_n_tested(any_pos_n, any_tested_n)
  ) %>%
  dplyr::select(
    `Sample Type` = sample_type,
    Year,
    `Samples, n`,
    `Unique Units, n`,
    `Time-Known Subset, n (%)`,
    `Culture +, +/Tested (%)`,
    `PCR +, +/Tested (%)`,
    `Any Test +, +/Tested (%)`
  )

table1_out

###################################
# Generation of Supp Table 1 Data #
###################################

fmt_n_pct <- function(n, d) {
  ifelse(
    is.na(d) | d == 0,
    "",
    sprintf("%d (%.1f)", n, 100 * n / d)
  )
}

fmt_pos_tested <- function(pos, tested) {
  ifelse(
    is.na(tested) | tested == 0,
    "",
    sprintf("%d/%d (%.1f)", pos, tested, 100 * pos / tested)
  )
}

fmt_n_units <- function(n, units) {
  ifelse(
    is.na(n),
    "",
    sprintf("%d (%d)", n, units)
  )
}

#Prep Data#

supp__soil_dat <- dat %>%
  mutate(
    sample_year   = as.character(sample_year),
    distance_m    = as.numeric(distance_m),
    distance_lab  = paste0(as.integer(distance_m), " m"),
    time_known    = !is.na(year_of_death),
    carcass_id    = as.character(carcass_id),
    site_id       = as.character(site_id),
    unique_id     = as.character(unique_id)
  ) %>%
  filter(
    substrate == "Soil",
    carcass_association == "Y",
    distance_m %in% c(0, 1, 3, 5, 15)
  ) %>%
  filter(
    !str_detect(carcass_id, "^WOLF-23-01$"),
    !str_detect(site_id, "^WOLF-23-01$"),
    !str_detect(unique_id, "^WOLF-23-01")
  )

#Summarize by Distance-Year#

soil_years <- supp__soil_dat %>%
  group_by(distance_lab, sample_year) %>%
  summarise(
    samples_n       = n(),
    unique_units_n  = n_distinct(site_id),
    time_known_n    = sum(time_known, na.rm = TRUE),
    culture_pos_n   = sum(culture_ac_pos == "Y", na.rm = TRUE),
    culture_tested_n= sum(culture_done == "Y", na.rm = TRUE),
    pcr_pos_n       = sum(pcr_ac_pos == "Y", na.rm = TRUE),
    pcr_tested_n    = sum(pcr_done == "Y", na.rm = TRUE),
    any_pos_n       = sum(ac_any_pos == "Y", na.rm = TRUE),
    .groups = "drop"
  )

soil_totals <- supp__soil_dat %>%
  group_by(distance_lab) %>%
  summarise(
    sample_year      = "TOTAL",
    samples_n        = n(),
    unique_units_n   = n_distinct(site_id),
    time_known_n     = sum(time_known, na.rm = TRUE),
    culture_pos_n    = sum(culture_ac_pos == "Y", na.rm = TRUE),
    culture_tested_n = sum(culture_done == "Y", na.rm = TRUE),
    pcr_pos_n        = sum(pcr_ac_pos == "Y", na.rm = TRUE),
    pcr_tested_n     = sum(pcr_done == "Y", na.rm = TRUE),
    any_pos_n        = sum(ac_any_pos == "Y", na.rm = TRUE),
    .groups = "drop"
  )

soil_stats <- bind_rows(soil_years, soil_totals)

#Build Table Shell#

table_shell <- tibble(
  distance_lab = c(
    "0 m", "0 m", "0 m", "0 m",
    "1 m", "1 m", "1 m",
    "3 m", "3 m", "3 m",
    "5 m", "5 m", "5 m",
    "15 m", "15 m", "15 m"
  ),
  sample_year = c(
    "2022", "2023", "2024", "TOTAL",
    "2023", "2024", "TOTAL",
    "2023", "2024", "TOTAL",
    "2023", "2024", "TOTAL",
    "2023", "2024", "TOTAL"
  )
)

supp_table_soil_distance <- table_shell %>%
  left_join(soil_stats, by = c("distance_lab", "sample_year")) %>%
  mutate(
    `Sampling Distance`            = distance_lab,
    Year                           = sample_year,
    `Samples, n (Unique Units*)`   = fmt_n_units(samples_n, unique_units_n),
    `Known Time of Death*, n (%)`  = fmt_n_pct(time_known_n, samples_n),
    `Culture +/Tested (%)`         = fmt_pos_tested(culture_pos_n, culture_tested_n),
    `PCR +/Tested (%)`             = fmt_pos_tested(pcr_pos_n, pcr_tested_n),
    `Any Test +/Tested (%)`        = fmt_pos_tested(any_pos_n, samples_n)
  ) %>%
  dplyr::select(
    `Sampling Distance`,
    Year,
    `Samples, n (Unique Units*)`,
    `Known Time of Death*, n (%)`,
    `Culture +/Tested (%)`,
    `PCR +/Tested (%)`,
    `Any Test +/Tested (%)`
  )

supp_table_soil_distance

```

## Interval Censoring Setup & Turnbull Curve Construction
```{r Survival Analysis: Interval Censoring Setup & Turnbull Curve Construction, cache=FALSE}
# Build interval-censored bounds per carcass_id × substrate #
visits <- visits %>%
  arrange(carcass_id, substrate, time_since_death_years)

paired_ids <- visits %>%
  distinct(carcass_id, substrate) %>%
  count(carcass_id, name = "n_sub") %>%
  filter(n_sub == 2) %>%
  pull(carcass_id)

visits_paired <- visits %>%
  filter(carcass_id %in% paired_ids) %>%
  arrange(carcass_id, substrate, time_since_death_years)

# adjust carcasses with tissue neg-->pos--neg recurrence #
tissue_force_L0_ids <- c("MX-21-04", "MX-21-16", "MX-21-26A")

ic_tsd <- visits_paired %>%
  group_by(carcass_id, substrate) %>%
  summarise(
    first_t = min(time_since_death_years, na.rm = TRUE),
    last_t  = max(time_since_death_years, na.rm = TRUE),

    any_pos = any(pos),
    last_status = dplyr::last(pos),

    first_pos_t = if (any(pos)) min(time_since_death_years[pos], na.rm = TRUE) else NA_integer_,
    last_pos_t = if (any(pos)) max(time_since_death_years[pos], na.rm = TRUE) else NA_integer_,

    first_neg_after_last_pos_t = if (any(pos)) {
      lp <- max(time_since_death_years[pos], na.rm = TRUE)
      cand <- time_since_death_years[!pos & time_since_death_years > lp]
      if (length(cand) > 0) min(cand, na.rm = TRUE) else NA_integer_
    } else NA_integer_,

    censor_type = case_when(
      !any_pos ~ "left",
      last_status ~ "right",
      TRUE ~ "interval"
    ),

    L = case_when(
      censor_type == "left" ~ NA_real_,
      TRUE ~ as.numeric(last_pos_t)
    ),
    R = case_when(
      censor_type == "left" ~ as.numeric(first_t),
      censor_type == "right" ~ NA_real_,
      TRUE ~ as.numeric(first_neg_after_last_pos_t)
    ),
    .groups = "drop"
  )

# interval2 bounds for Surv(L2, R2, type="interval2") #
min_tsd <- min(ic_tsd$R, na.rm = TRUE)

curve_known_tsd2 <- ic_tsd %>%
  mutate(
    L2 = if_else(censor_type == "left",  as.numeric(min_tsd - 1), as.numeric(L)),
    R2 = if_else(censor_type == "right", Inf,                    as.numeric(R)),
    substrate = factor(substrate, levels = c("Tissue","Soil"))
  ) %>%
  filter(!is.na(L2), !is.na(R2), L2 <= R2)

# Curves #
fit_known_tsd <- survfit(Surv(L2, R2, type = "interval2") ~ substrate,
                         data = curve_known_tsd2)

plot(fit_known_tsd,
     xlab = "Years since death",
     ylab = "Pr(still culture positive)",
     conf.int = TRUE,
     mark.time = TRUE,
     main = "Probability of culture positivity following death")
legend("topright", legend = levels(curve_known_tsd2$substrate), lty = 1:2, bty = "n")

summary(fit_known_tsd)$table

# Nicer Plot #

theme_pub <- theme_classic(base_size = 11) +
  theme(
    plot.title      = element_text(face = "bold", size = 12, margin = margin(b = 4)),
    plot.subtitle   = element_text(size = 10, margin = margin(b = 6)),
    axis.title      = element_text(face = "bold"),
    axis.text       = element_text(color = "black"),
    panel.border    = element_rect(fill = NA, color = "black", linewidth = 0.4),
    strip.background= element_rect(fill = "grey95", color = NA),
    strip.text      = element_text(face = "bold"),
    panel.spacing   = unit(0.9, "lines"),
    plot.caption    = element_text(size = 8, hjust = 0, margin = margin(t = 6))
  )

tidy_survfit <- function(fit) {
  s <- summary(fit)

  tibble(
    time  = s$time,
    surv  = s$surv,
    lower = s$lower,
    upper = s$upper,
    strata = if (!is.null(s$strata)) as.character(s$strata) else "All"
  ) %>%
    mutate(group = str_replace(strata, ".*=", "")) %>%
    filter(is.finite(time)) %>%              # drop Inf if present
    arrange(group, time)
}

plot_survfit_gg <- function(fit, xlab, title, x_breaks = NULL, x_limits = NULL) {
  df <- tidy_survfit(fit)

  if (is.null(x_breaks)) {
    x_breaks <- pretty(range(df$time, na.rm = TRUE), n = 6)
  }

  ggplot(df, aes(x = time, y = surv, color = group, linetype = group)) +
    geom_step(linewidth = 1.2) +
    # CI as dashed steps (clean + journal-friendly)
    geom_step(aes(y = lower), linewidth = 0.6, alpha = 0.6) +
    geom_step(aes(y = upper), linewidth = 0.6, alpha = 0.6) +
    scale_y_continuous(limits = c(0, 1), expand = c(0, 0)) +
    scale_x_continuous(breaks = x_breaks, limits = x_limits) +
    labs(
      x = xlab,
      y = "Pr(still culture positive)",
      title = title,
      color = NULL,
      linetype = NULL
    ) +
    theme_minimal(base_size = 12) +
    theme(
      panel.grid.minor = element_blank(),
      legend.position = "top",
      plot.title = element_text(face = "bold")
    )
}

x_breaks_tsd <- seq(floor(min(curve_known_tsd2$L2, na.rm = TRUE)),
                    ceiling(max(curve_known_tsd2$R2[is.finite(curve_known_tsd2$R2)], na.rm = TRUE)),
                    by = 1)

p_known <- plot_survfit_gg(
  fit_known_tsd,
  xlab = "Years Since Death",
  title = NULL,
  x_breaks = x_breaks_tsd
) + theme_pub
p_known

```

## Log-rank & Paired Permutation Tests
```{r Survival Analysis: Log-rank & Paired Permutation Tests, cache=FALSE}
# Log-rank analogue #
curve_known_tsd2 <- curve_known_tsd2 %>%
  mutate(substrate = factor(substrate))

test_ic_logrank <- interval::ictest(
  Surv(L2, R2, type = "interval2") ~ substrate,
  data = curve_known_tsd2,
  scores = "logrank1",
  method = "pclt"
)

test_ic_logrank

# Paired Permutation Log-rank analogue #
extract_Z_safe <- function(obj) {
  z <- NULL
  if (!is.null(obj$Z)) z <- obj$Z
  else if (!is.null(obj$statistic)) z <- obj$statistic
  z <- suppressWarnings(as.numeric(z))
  if (length(z) == 0) return(NA_real_)
  z[1]
}

paired_perm_ictest <- function(dat, B = 5000, seed = 1) {
  set.seed(seed)

  dat <- dat %>%
    mutate(substrate = factor(substrate)) %>%
    arrange(carcass_id, substrate)

  tab <- table(dat$carcass_id)
  if (any(tab != 2)) {
    stop("Not all carcass_id have exactly 2 rows (Soil + Tissue). Check pairing filter.")
  }

  obs <- interval::ictest(Surv(L2, R2, type="interval2") ~ substrate,
                          data = dat, scores="logrank1", method="pclt")
  obs_z <- extract_Z_safe(obs)

  ids <- unique(dat$carcass_id)

  perm_z <- vapply(seq_len(B), function(i) {
    flip_ids <- ids[runif(length(ids)) < 0.5]

    datp <- dat
    idx <- datp$carcass_id %in% flip_ids

    s <- as.character(datp$substrate)
    s[idx] <- ifelse(s[idx] == "Soil", "Tissue",
              ifelse(s[idx] == "Tissue", "Soil", s[idx]))
    datp$substrate <- factor(s, levels = levels(dat$substrate))

    z <- tryCatch(
      extract_Z_safe(interval::ictest(Surv(L2, R2, type="interval2") ~ substrate,
                                      data = datp, scores="logrank1", method="pclt")),
      error = function(e) NA_real_
    )
    z
  }, numeric(1))

  keep <- !is.na(perm_z)
  perm_z2 <- perm_z[keep]

  p_two_sided <- (sum(abs(perm_z2) >= abs(obs_z)) + 1) / (length(perm_z2) + 1)

  list(
    observed_test = obs,
    observed_Z = obs_z,
    perm_p_two_sided = p_two_sided,
    B_requested = B,
    B_used = length(perm_z2),
    n_failed = sum(!keep)
  )
}

perm_res2 <- paired_perm_ictest(curve_known_tsd2, B = 5000, seed = 1)
perm_res2$observed_test
perm_res2$perm_p_two_sided
```

## Semi-Parametric Regression Models
```{r, Survival Analysis: Semi-Parametric Regression Models, cache=TRUE}

library(tidyverse)
library(survival)
library(icenReg)
library(foreach)
library(doParallel)

# Prep: filter data & define “pos”, keep paired carcasses , collapse habitat classes#

dat0 <- dat %>%
  filter(
    carcass_association == "Y",
    !is.na(year_of_death),
    site_id != "WOLF-23-01",
    carcass_id != "WOLF-23-01",
    substrate %in% c("Soil","Tissue"),
    !is.na(culture_ac_pos),
    !is.na(time_since_death_years)
  ) %>%
  mutate(pos = culture_ac_pos == "Y",
  time_since_death_years = as.numeric(time_since_death_years))

curve_ids <- dat0 %>%
  distinct(carcass_id, substrate) %>%
  count(carcass_id, name = "n_sub") %>%
  filter(n_sub == 2) %>%
  pull(carcass_id)

dist_keep <- c(0, 1, 3, 5)

soil_dat <- dat0 %>%
  filter(
    substrate == "Soil",
    carcass_id %in% curve_ids,
    !is.na(distance_m),
    as.numeric(distance_m) %in% dist_keep) %>%
  mutate(distance_m = as.numeric(distance_m))

  #3-level habitat classes by moisture#
make_habitat_moist3 <- function(x) {
  x <- as.character(x)
  out <- dplyr::case_when(
    x == "Tundra Barrens" ~ "Dry",
    x %in% c("Hummock Tundra", "Tundra Meadow") ~ "Medium",
    x %in% c("Riparian Tundra", "Riverbed", "Tundra Wetland") ~ "Wet",
    TRUE ~ NA_character_
  )

  factor(
    out,
    levels = c(
      "Dry",
      "Medium",
      "Wet"
    )
  )
}

# Build interval-censored bounds #

make_ic_series <- function(dat_series, series_vars) {

  visits <- dat_series %>%
    group_by(across(all_of(series_vars)), time_since_death_years) %>%
    summarise(
      pos = any(pos, na.rm = TRUE),

      carcass_size   = dplyr::first(na.omit(carcass_size)),
      habitat_class  = dplyr::first(na.omit(habitat_class)),
      year_of_death         = dplyr::first(na.omit(year_of_death)),
      sample_year    = suppressWarnings(min(sample_year, na.rm = TRUE)),

      .groups = "drop"
    ) %>%
    arrange(across(all_of(series_vars)), time_since_death_years)

  ic <- visits %>%
    group_by(across(all_of(series_vars))) %>%
    summarise(
      first_t = min(time_since_death_years, na.rm = TRUE),

      any_pos = any(pos),
      last_status = dplyr::last(pos),

      last_pos_t = if (any(pos)) max(time_since_death_years[pos], na.rm = TRUE) else NA_real_,

      first_neg_after_last_pos_t = if (any(pos)) {
        lp <- max(time_since_death_years[pos], na.rm = TRUE)
        cand <- time_since_death_years[!pos & time_since_death_years > lp]
        if (length(cand) > 0) min(cand, na.rm = TRUE) else NA_real_
      } else NA_real_,

      censor_type = case_when(
        !any_pos ~ "left",
        last_status ~ "right",
        TRUE ~ "interval"
      ),

      L = case_when(
        censor_type == "left" ~ NA_real_,
        TRUE ~ last_pos_t
      ),
      R = case_when(
        censor_type == "left"  ~ first_t,
        censor_type == "right" ~ NA_real_,
        TRUE ~ first_neg_after_last_pos_t
      ),

      carcass_size  = dplyr::first(na.omit(carcass_size)),
      habitat_class = dplyr::first(na.omit(habitat_class)),
      year_of_death        = dplyr::first(na.omit(year_of_death)),
      sample_year   = dplyr::first(na.omit(sample_year)),

      .groups = "drop"
    )

  min_R <- min(ic$R, na.rm = TRUE)

  ic2 <- ic %>%
    mutate(
      L2 = if_else(censor_type == "left",  as.numeric(min_R - 1), as.numeric(L)),
      R2 = if_else(censor_type == "right", Inf,                    as.numeric(R))
    ) %>%
    filter(!is.na(L2), !is.na(R2), L2 <= R2)

  list(visits = visits, ic = ic2)
}

### TISSUE MODELS ###

set.seed(1)

tissue_dat <- dat0 %>%
  filter(
    substrate == "Tissue",
    carcass_id %in% curve_ids
  )

tissue_obj <- make_ic_series(tissue_dat, series_vars = c("carcass_id"))

ic_tissue_raw <- tissue_obj$ic %>%
  transmute(
    carcass_id,
    L2, R2, censor_type,
    carcass_size  = factor(carcass_size),
    habitat_class = factor(habitat_class),
    habitat_moist3 = make_habitat_moist3(habitat_class),
    year          = as.numeric(year_of_death),
    year_num      = as.numeric(year_of_death) - 2021
  ) %>%
  filter(
    !is.na(L2), !is.na(R2),
    !is.na(carcass_size),
    !is.na(habitat_moist3),
    !is.na(year),
    L2 <= R2
  )

stopifnot(nrow(ic_tissue_raw) == 37)

ic_tissue_red <- ic_tissue_raw %>%
  mutate(
    year_bin = ifelse(year <= 2021, "2021-or-earlier", "2022-or-later"),
    year_bin = factor(year_bin, levels = c("2021-or-earlier", "2022-or-later"))
  )

ic_tissue_red <- ic_tissue_red %>%
  mutate(
    carcass_size = relevel(carcass_size, ref = levels(carcass_size)[1]))

ctrl <- makeCtrls_icsp(useGA = FALSE, maxIter = 30000, baseUpdates = 10)

fit_icsp_ci <- function(formula, data, bs = 500, useMCores = TRUE, ctrls = NULL) {

  if (is.null(ctrls)) {
    ctrls <- icenReg::makeCtrls_icsp(useGA = FALSE, maxIter = 30000, baseUpdates = 10)
  }

  fit <- icenReg::ic_sp(
    formula,
    data       = data,
    model      = "ph",
    B          = c(0, 1),
    bs_samples = bs,
    useMCores  = useMCores,
    controls   = ctrls
  )

  b0 <- coef(fit)
  bsMat <- fit$bsMat
  bs_ok <- bsMat[stats::complete.cases(bsMat), , drop = FALSE]

  ci_log <- t(apply(bs_ok, 2, stats::quantile, probs = c(0.025, 0.975), na.rm = TRUE))

  tab <- tibble::tibble(
    term  = names(b0),
    logHR = as.numeric(b0),
    HR    = exp(logHR),
    CI_lo = exp(ci_log[, 1]),
    CI_hi = exp(ci_log[, 2])
  )

  list(fit = fit, table = tab, n_boot_used = nrow(bs_ok))
}

#Carcass Size Univariable Model#
m_size <- fit_icsp_ci(
  Surv(L2, R2, type = "interval2") ~ carcass_size,
  data = ic_tissue_raw,
  bs   = 500,
  useMCores = TRUE,
  ctrls = ctrl
)

m_size$table

#Habitat Class Univariable Model#
m_hab_red <- fit_icsp_ci(
  Surv(L2, R2, type="interval2") ~ habitat_moist3,
  data = ic_tissue_red,
  bs = 500,
  useMCores = TRUE,
  ctrls = ctrl
)

m_hab_red$table

#Year of Death (continuous) Univariable Model#
m_year_cont <- fit_icsp_ci(
  Surv(L2, R2, type = "interval2") ~ year_num,
  data = ic_tissue_raw,
  bs   = 500,
  useMCores = TRUE,
  ctrls = ctrl
)

m_year_cont$table

#Year of Death (categorical) Univariable Model#
m_year_bin <- fit_icsp_ci(
  Surv(L2, R2, type="interval2") ~ year_bin,
  data = ic_tissue_red,
  bs = 500,
  useMCores = TRUE,
  ctrls = ctrl
)

m_year_bin$table

#Carcass Size and Year of Death (continuous) Model#
m_size_year <- fit_icsp_ci(
  Surv(L2, R2, type="interval2") ~ carcass_size + year_num,
  data = ic_tissue_raw,
  bs = 500,
  useMCores = TRUE,
  ctrls = ctrl
)

m_size_year$table

#Habitat Class and Year of Death (continuous) Model#
m_hab_year <- fit_icsp_ci(
  Surv(L2, R2, type="interval2") ~ habitat_moist3 + year_num,
  data = ic_tissue_red,
  bs = 500,
  useMCores = TRUE,
  ctrls = ctrl
)

m_hab_year$table

#Habitat Class, Carcass Size, and Year of Death (continuous) Model#
m_full_tiss <- fit_icsp_ci(
  Surv(L2, R2, type="interval2") ~ habitat_moist3 + year_num + carcass_size,
  data = ic_tissue_red,
  bs = 500,
  useMCores = TRUE,
  ctrls = ctrl
)

m_full_tiss$table

### SOIL MODELS ###

soil_obj <- make_ic_series(soil_dat, series_vars = c("carcass_id", "distance_m"))

ic_soil_raw <- soil_obj$ic %>%
  transmute(
    carcass_id,
    distance_m,
    L2, R2, censor_type,
    carcass_size  = factor(carcass_size),
    habitat_class = factor(habitat_class),
    habitat_moist3 = make_habitat_moist3(habitat_class),
    year          = as.numeric(year_of_death),
    year_num      = as.numeric(year_of_death) - 2021
  ) %>%
  filter(
    !is.na(L2), !is.na(R2),
    !is.na(distance_m),
    !is.na(carcass_size),
    !is.na(habitat_moist3),
    !is.na(year_num),
    L2 <= R2
  ) %>%
  mutate(
    distance_m = as.numeric(distance_m),
    distance_f = factor(distance_m, levels = dist_keep),
    distance_f = forcats::fct_relevel(distance_f, "0"),
    carcass_size = relevel(carcass_size, ref = levels(carcass_size)[1]),
    habitat_class = droplevels(habitat_class)
  )

ic_soil_cont <- ic_soil_raw %>%
  mutate(
    dist_m = as.numeric(distance_m),
    year_c = year_num - mean(year_num, na.rm = TRUE)
  )

#Carcass-Level Cluster Bootstrap#
cluster_boot_icsp <- function(dat_ic, formula, idvar = "carcass_id",
                              B = 500, seed = 1, ctrls = ctrl) {

  set.seed(seed)
  ids <- unique(dat_ic[[idvar]])

  # point estimate
  fit0 <- icenReg::ic_sp(formula, data = dat_ic, model = "ph", B = c(0,1), bs_samples = 0, controls = ctrls)
  b0 <- coef(fit0)

  boot <- matrix(NA_real_, nrow = B, ncol = length(b0))
  colnames(boot) <- names(b0)
  n_fail <- 0L

  for (b in seq_len(B)) {
    samp_ids <- sample(ids, size = length(ids), replace = TRUE)

    boot_dat <- purrr::map_dfr(samp_ids, ~ dplyr::filter(dat_ic, .data[[idvar]] == .x))

    fit_b <- tryCatch(
      icenReg::ic_sp(formula, data = boot_dat, model = "ph", B = c(0,1), bs_samples = 0, controls = ctrls),
      error = function(e) NULL
    )

    if (is.null(fit_b)) { n_fail <- n_fail + 1L; next }

    cb <- tryCatch(coef(fit_b), error = function(e) NULL)
    if (is.null(cb)) { n_fail <- n_fail + 1L; next }

    boot[b, names(cb)] <- as.numeric(cb)
  }

  boot_df <- as_tibble(boot) %>% dplyr::filter(if_all(everything(), ~ !is.na(.x)))

  ci_log <- t(apply(boot_df, 2, quantile, probs = c(0.025, 0.975), na.rm = TRUE))

  tab <- tibble(
    term  = names(b0),
    logHR = as.numeric(b0),
    HR    = exp(logHR),
    CI_lo = exp(ci_log[, 1]),
    CI_hi = exp(ci_log[, 2])
  )

  list(
    fit0 = fit0,
    table = tab,
    n_used = nrow(boot_df),
    n_failed = n_fail
  )
}

#Carcass Level Warm Cluster Bootstrap#
cluster_boot_icsp_warm <- function(dat_ic, formula, idvar="carcass_id",
                                   B = 300, seed = 1, ctrls = ctrl_mv_warm) {
  set.seed(seed)
  ids <- unique(dat_ic[[idvar]])

  fit0 <- icenReg::ic_sp(formula, data = dat_ic, model="ph", B=c(0,1), bs_samples=0, controls=ctrls)
  b0 <- coef(fit0)

  boot <- matrix(NA_real_, nrow = B, ncol = length(b0))
  colnames(boot) <- names(b0)

  n_fail <- 0L

  for (b in seq_len(B)) {
    samp_ids <- sample(ids, length(ids), replace = TRUE)
    boot_dat <- purrr::map_dfr(samp_ids, ~ dplyr::filter(dat_ic, .data[[idvar]] == .x))

    fit_b <- tryCatch(
      icenReg::ic_sp(formula, data = boot_dat, model="ph", B=c(0,1), bs_samples=0, controls=ctrls),
      error = function(e) NULL
    )

    if (is.null(fit_b)) { n_fail <- n_fail + 1L; next }

    cb <- tryCatch(coef(fit_b), error = function(e) NULL)
    if (is.null(cb)) { n_fail <- n_fail + 1L; next }

    boot[b, names(cb)] <- as.numeric(cb)
  }

  boot_ok <- boot[stats::complete.cases(boot), , drop=FALSE]

  ci_log <- t(apply(boot_ok, 2, stats::quantile, probs=c(0.025,0.975), na.rm=TRUE))
  tab <- tibble::tibble(
    term  = names(b0),
    logHR = as.numeric(b0),
    HR    = exp(logHR),
    CI_lo = exp(ci_log[,1]),
    CI_hi = exp(ci_log[,2])
  )

  list(fit0 = fit0, table = tab, n_used = nrow(boot_ok), n_failed = n_fail)
}

#Distance (factor) Univariable Model#
m_dist <- cluster_boot_icsp(
  dat_ic  = ic_soil_raw,
  formula = Surv(L2, R2, type="interval2") ~ distance_f,
  idvar   = "carcass_id",
  B       = 500,
  seed    = 1,
  ctrls   = ctrl
)

m_dist$table
m_dist$n_used
m_dist$n_failed

#Distance (continuous) Univariable Model#
m_dist_cont <- cluster_boot_icsp(
  dat_ic  = ic_soil_cont,
  formula = Surv(L2, R2, type="interval2") ~ dist_m,
  idvar   = "carcass_id",
  B       = 500,
  seed    = 1,
  ctrls   = ctrl
)

m_dist_cont$table

#Carcass Size Univariable Model#
m_size_soil <- cluster_boot_icsp(
  ic_soil_raw,
  Surv(L2, R2, type="interval2") ~ carcass_size,
  idvar = "carcass_id",
  B = 500,
  seed = 1,
  ctrls = ctrl
)

m_size_soil$table

#Year of Death (continuous) Univariable Model#
m_year_soil <- cluster_boot_icsp(
  ic_soil_raw,
  Surv(L2, R2, type="interval2") ~ year_num,
  idvar = "carcass_id",
  B = 500, 
  seed = 1, 
  ctrls = ctrl
)

m_year_soil$table

#Habitat Class Univariable Model#
m_hab_soil <- cluster_boot_icsp(
  ic_soil_raw,
  Surv(L2, R2, type="interval2") ~ habitat_moist3,
  idvar = "carcass_id",
  B = 500, 
  seed = 1, 
  ctrls = ctrl
)

m_hab_soil$table

#Distance (factor) and Year of Death (continuous) Multivariable Model#
ctrl_base <- icenReg::makeCtrls_icsp(useGA = TRUE, maxIter = 30000, baseUpdates = 5)

ic_soil_mv <- ic_soil_raw %>%
  dplyr::mutate(
    year_c = year_num - mean(year_num, na.rm = TRUE) #centered
  )

fit_full <- icenReg::ic_sp(
  Surv(L2, R2, type="interval2") ~ distance_f + year_c,
  data = ic_soil_mv,
  model = "ph", B = c(0,1),
  bs_samples = 0,
  controls = ctrl_base
)

ctrl_mv_warm <- icenReg::makeCtrls_icsp(
  useGA = TRUE,
  maxIter = 30000,
  baseUpdates = 5,
  regStart = coef(fit_full)
)

m_dist_year <- cluster_boot_icsp_warm(
  dat_ic  = ic_soil_mv,
  formula = Surv(L2, R2, type="interval2") ~ distance_f + year_c,
  idvar   = "carcass_id",
  B       = 200,
  seed    = 1,
  ctrls   = ctrl_mv_warm
)

m_dist_year$table

#Distance (continuous) and Year of Death (continuous) Sensitivity Model#
fit_full_cont <- icenReg::ic_sp(
  Surv(L2, R2, type = "interval2") ~ dist_m + year_c,
  data       = ic_soil_cont,
  model      = "ph",
  B          = c(0,1),
  bs_samples = 0,
  controls   = ctrl_base
)

ctrl_warm_cont <- icenReg::makeCtrls_icsp(
  useGA = TRUE,
  maxIter = 30000,
  baseUpdates = 5,
  regStart = coef(fit_full_cont)
)

m_dist_year_cont <- cluster_boot_icsp_warm(
  dat_ic  = ic_soil_cont,
  formula = Surv(L2, R2, type = "interval2") ~ dist_m + year_c,
  idvar   = "carcass_id",
  B       = 200,
  seed    = 1,
  ctrls   = ctrl_warm_cont
)

m_dist_year_cont$table

#Distance (factor), Carcass Size, and Year of Death (continuous) Multivariable Model#
fit_full_size <- icenReg::ic_sp(
  Surv(L2, R2, type="interval2") ~ distance_f + year_c + carcass_size,
  data       = ic_soil_mv,
  model      = "ph",
  B          = c(0,1),
  bs_samples = 0,
  controls   = ctrl_base
)

ctrl_warm_size <- icenReg::makeCtrls_icsp(
  useGA = TRUE, maxIter = 30000, baseUpdates = 5,
  regStart = coef(fit_full_size)
)

m_dist_size_year <- cluster_boot_icsp_warm(
  dat_ic  = ic_soil_mv,
  formula = Surv(L2, R2, type="interval2") ~ distance_f + year_c + carcass_size,
  idvar   = "carcass_id",
  B       = 200,
  seed    = 1,
  ctrls   = ctrl_warm_size
)

m_dist_size_year$table

#Distance (continuous), Carcass Size, and Year of Death (continuous) Sensitivity Model#
fit_full_cont_size <- icenReg::ic_sp(
  Surv(L2, R2, type = "interval2") ~ dist_m + year_c + carcass_size,
  data       = ic_soil_cont,
  model      = "ph",
  B          = c(0,1),
  bs_samples = 0,
  controls   = ctrl_base
)

ctrl_warm_cont_size <- icenReg::makeCtrls_icsp(
  useGA = TRUE,
  maxIter = 30000,
  baseUpdates = 5,
  regStart = coef(fit_full_cont_size)
)

m_dist_size_year_cont <- cluster_boot_icsp_warm(
  dat_ic  = ic_soil_cont,
  formula = Surv(L2, R2, type = "interval2") ~ dist_m + year_c + carcass_size,
  idvar   = "carcass_id",
  B       = 200,
  seed    = 1,
  ctrls   = ctrl_warm_cont_size
)

m_dist_size_year_cont$table

#Distance (factor), Habitat Class, and Year of Death (continuous) Multivariable Model# 
fit_full_hab <- icenReg::ic_sp(
  Surv(L2, R2, type="interval2") ~ distance_f + year_c + habitat_moist3,
  data = ic_soil_mv,
  model = "ph", B = c(0,1),
  bs_samples = 0,
  controls = ctrl_base
)

ctrl_warm_hab <- icenReg::makeCtrls_icsp(
  useGA = TRUE, maxIter = 30000, baseUpdates = 5,
  regStart = coef(fit_full_hab)
)

m_dist_hab_year <- cluster_boot_icsp_warm(
  ic_soil_mv,
  Surv(L2, R2, type="interval2") ~ distance_f + year_c + habitat_moist3,
  idvar = "carcass_id",
  B = 200, seed = 1,
  ctrls = ctrl_warm_hab
)

m_dist_hab_year$table

#Distance (continuous), Habitat Class, and Year of Death (continuous) Sensitivity Model# 
fit_full_cont_hab <- icenReg::ic_sp(
  Surv(L2, R2, type = "interval2") ~ dist_m + year_c + habitat_moist3,
  data       = ic_soil_cont,
  model      = "ph",
  B          = c(0,1),
  bs_samples = 0,
  controls   = ctrl_base
)

ctrl_warm_cont_hab <- icenReg::makeCtrls_icsp(
  useGA = TRUE,
  maxIter = 30000,
  baseUpdates = 5,
  regStart = coef(fit_full_cont_hab)
)

m_dist_hab_year_cont <- cluster_boot_icsp_warm(
  dat_ic  = ic_soil_cont,
  formula = Surv(L2, R2, type = "interval2") ~ dist_m + year_c + habitat_moist3,
  idvar   = "carcass_id",
  B       = 200,
  seed    = 1,
  ctrls   = ctrl_warm_cont_hab
)

m_dist_hab_year_cont$table

#Distance (factor), Habitat Class, Carcass Size, and Year of Death (continuous) Multivariable Model# 
fit_full_soil <- icenReg::ic_sp(
  Surv(L2, R2, type = "interval2") ~ distance_f + year_c + habitat_moist3 + carcass_size,
  data       = ic_soil_mv,
  model      = "ph",
  B          = c(0,1),
  bs_samples = 0,
  controls   = ctrl_base
)

ctrl_warm_full_soil <- icenReg::makeCtrls_icsp(
  useGA = TRUE,
  maxIter = 30000,
  baseUpdates = 5,
  regStart = coef(fit_full_soil)
)

m_full_soil <- cluster_boot_icsp_warm(
  dat_ic  = ic_soil_mv,
  formula = Surv(L2, R2, type = "interval2") ~ distance_f + year_c + habitat_moist3 + carcass_size,
  idvar   = "carcass_id",
  B       = 200,
  seed    = 1,
  ctrls   = ctrl_warm_full_soil
)

m_full_soil$table

```


# Latent Bacterial Population Modelling
## Setup & Model Fitting
```{r Latent Bacterial Population Modelling, cache=FALSE}

#Data Setup#
dat0 <- dat %>%
  filter(
    carcass_association == "Y",
    substrate %in% c("Tissue", "Soil"),
    !is.na(year_of_death),
    !is.na(carcass_id),
    carcass_id != "WOLF-23-01",
    culture_done == "Y",
    culture_ac_pos %in% c("Y", "N")
  ) %>%
  mutate(
    time_since_death_years = as.numeric(time_since_death_years),
    distance_m = as.numeric(distance_m),
    year_num = as.numeric(year_of_death) - 2021,
    carcass_size = factor(carcass_size),
    habitat_moist3 = make_habitat_moist3(habitat_class)
  )

paired_ids <- dat0 %>%
  group_by(carcass_id) %>%
  summarise(n_sub = n_distinct(substrate), .groups = "drop") %>%
  filter(n_sub == 2) %>%
  pull(carcass_id)

dat0 <- dat0 %>% filter(carcass_id %in% paired_ids)

stopifnot(!any(is.na(dat0$habitat_moist3)))

#Collapse Data to Visits (by substrate)#
tiss_vis <- dat0 %>%
  filter(substrate == "Tissue") %>%
  group_by(carcass_id, time_since_death_years) %>%
  summarise(
    y = as.integer(any(culture_ac_pos == "Y")),
    year_num = first(year_num),
    carcass_size = first(carcass_size),
    habitat_class = first(habitat_class),
    habitat_moist3 = first(habitat_moist3),
    .groups = "drop"
  )

soil_vis <- dat0 %>%
  filter(substrate == "Soil",
         !is.na(distance_m),
         as.numeric(distance_m) %in% dist_keep) %>%
  mutate(
    distance_m = as.numeric(distance_m),
    distance_f = factor(distance_m, levels = dist_keep),
    distance_f = forcats::fct_relevel(distance_f, "0")
  ) %>%
  group_by(carcass_id, distance_f, time_since_death_years) %>%
  summarise(
    y = as.integer(any(culture_ac_pos == "Y")),
    year_num = first(year_num),
    carcass_size = first(carcass_size),
    habitat_moist3 = first(habitat_moist3),
    distance_m = first(distance_m),
    .groups = "drop"
  )

soil_vis_cont <- dat0 %>%
  filter(
    substrate == "Soil",
    !is.na(distance_m),
    as.numeric(distance_m) %in% dist_keep
  ) %>%
  mutate(distance_m = as.numeric(distance_m)) %>%
  group_by(carcass_id, distance_m, time_since_death_years) %>%
  summarise(
    y = as.integer(any(culture_ac_pos == "Y")),
    year_num = first(year_num),
    carcass_size = first(carcass_size),
    habitat_moist3 = first(habitat_moist3),
    .groups = "drop"
  )

#Fit cloglog Mechanistic Models#
library(glmmTMB)

m_tiss <- glmmTMB(
  y ~ time_since_death_years + habitat_moist3 + carcass_size + (1 | carcass_id),
  family = binomial(link = "cloglog"),
  data = tiss_vis
)

m_soil <- glmmTMB(
  y ~ time_since_death_years + distance_f + habitat_moist3 + carcass_size + (1 | carcass_id),
  family = binomial(link = "cloglog"),
  data = soil_vis
)

m_soil_cont <- glmmTMB(
  y ~ time_since_death_years + distance_m + habitat_moist3 + carcass_size + (1 | carcass_id),
  family = binomial(link = "cloglog"),
  data = soil_vis_cont
)

summary(m_tiss)
summary(m_soil)
summary(m_soil_cont)

b_time_tiss <- fixef(m_tiss)$cond["time_since_death_years"]
k_tiss <- -b_time_tiss
t_half_tiss <- log(2) / k_tiss
k_tiss; t_half_tiss

b_time_soil <- fixef(m_soil)$cond["time_since_death_years"]
k_soil <- -b_time_soil
t_half_soil <- log(2) / k_soil
k_soil; t_half_soil

b_dist <- fixef(m_soil)$cond[grep("^distance_f", names(fixef(m_soil)$cond))]
exp(b_dist)  # odds ratios / hazard-like contrasts vs 0m

b_dist_cont <- fixef(m_soil_cont)$cond["distance_m"]
att_per_m_cont <- exp(b_dist_cont)

c(att_per_m_cont = att_per_m_cont)

#Soil model with time*distance interaction term*
m_soil_timedistint <- glmmTMB(
  y ~ time_since_death_years * distance_f + habitat_moist3 + carcass_size + (1|carcass_id),
  family = binomial(link="cloglog"),
  data = soil_vis
)

summary(m_soil_timedistint) #muddied by sparsity of positives at distance --> retain model w/o interaction#

tidy_cloglog_exp <- function(mod) {
  sm <- summary(mod)$coefficients$cond
  est <- sm[, "Estimate"]
  se  <- sm[, "Std. Error"]
  z   <- sm[, "z value"]
  p   <- sm[, "Pr(>|z|)"]
  
  tibble(
    term = rownames(sm),
    estimate = est,
    std_error = se,
    z_value = z,
    p_value = p,
    exp_coef = exp(est),
    ci_low = exp(est - 1.96 * se),
    ci_high = exp(est + 1.96 * se)
  )
}

clog_tiss_tab <- tidy_cloglog_exp(m_tiss)
clog_soil_tab <- tidy_cloglog_exp(m_soil)
clog_soil_int_tab <- tidy_cloglog_exp(m_soil_timedistint)

clog_tiss_tab
clog_soil_tab
clog_soil_int_tab

```

## Predicted-Curve Plots
```{r Latent Bacterial Population Modelling: Predicted-Curve Plots, cache=FALSE}

library(tidyverse)
library(scales)

inv_cloglog <- function(eta) 1 - exp(-exp(eta))

pred_ci_cloglog <- function(model, newdata, level = 0.95) {
  pr <- predict(model, newdata = newdata, type = "link", se.fit = TRUE, re.form = NA)
  z <- qnorm(1 - (1 - level)/2)

  newdata %>%
    mutate(
      eta = pr$fit,
      se  = pr$se.fit,
      eta_lo = eta - z * se,
      eta_hi = eta + z * se,
      p      = inv_cloglog(eta),
      p_lo   = inv_cloglog(eta_lo),
      p_hi   = inv_cloglog(eta_hi)
    )
}

make_col <- function(mf, var, value, n) {
  if (is.factor(mf[[var]])) factor(value, levels = levels(mf[[var]])) else rep(as.character(value), n)
}

#Tissue Predicted Curve (ref = most common factor levels)#
mf <- model.frame(m_tiss)

ref_hab_t  <- names(sort(table(mf$habitat_moist3), decreasing = TRUE))[1]
ref_size_t <- names(sort(table(mf$carcass_size),  decreasing = TRUE))[1]

t_grid <- tibble(
  time_since_death_years = seq(min(mf$time_since_death_years, na.rm = TRUE),
                               max(mf$time_since_death_years, na.rm = TRUE),
                               length.out = 200),
  habitat_moist3 = make_col(mf, "habitat_moist3", ref_hab_t, 200),
  carcass_size = if (is.factor(mf$carcass_size)) {
    factor(ref_size_t, levels = levels(mf$carcass_size))
  } else {
    rep(ref_size_t, 200)
  }
)

t_pred <- pred_ci_cloglog(m_tiss, t_grid)

ggplot(t_pred, aes(time_since_death_years, p)) +
  geom_ribbon(aes(ymin = p_lo, ymax = p_hi), alpha = 0.2) +
  geom_line() +
  labs(
    x = "Years Since Death",
    y = "Predicted P(Culture+)",
    title = "Predicted Tissue Culture Positivity Over Time",
  ) +
  coord_cartesian(ylim = c(0, 1))

#Tissue Predicted Curve (ref = small carcass size)#
mf <- model.frame(m_tiss)

ref_hab_t <- names(sort(table(mf$habitat_moist3), decreasing = TRUE))[1]

size_want <- "1"
size_col <- if (is.factor(mf$carcass_size)) {
  factor(size_want, levels = levels(mf$carcass_size))
} else {
  size_want
}

t_grid_size1 <- tibble(
  time_since_death_years = seq(min(mf$time_since_death_years, na.rm = TRUE),
                               max(mf$time_since_death_years, na.rm = TRUE),
                               length.out = 200),
  habitat_moist3 = make_col(mf, "habitat_moist3", ref_hab_t, 200),
  carcass_size  = size_col
)

t_pred_size1 <- pred_ci_cloglog(m_tiss, t_grid_size1)

ggplot(t_pred_size1, aes(time_since_death_years, p)) +
  geom_ribbon(aes(ymin = p_lo, ymax = p_hi), alpha = 0.2) +
  geom_line() +
  labs(
    x = "Years since death",
    y = "Predicted P(culture+)",
    title = "Predicted Tissue Culture Positivity Over Time, Small Carcasses",
  ) +
  coord_cartesian(ylim = c(0, 1))

#Soil Predicted Curves by Distance (ref = most common factor levels)#

mf_s <- model.frame(m_soil)

ref_hab_s  <- names(sort(table(mf_s$habitat_moist3), decreasing = TRUE))[1]
ref_size_s <- names(sort(table(mf_s$carcass_size),  decreasing = TRUE))[1]

dist_levels <- levels(mf_s$distance_f)

t_seq <- seq(min(mf_s$time_since_death_years, na.rm = TRUE),
             max(mf_s$time_since_death_years, na.rm = TRUE),
             length.out = 200)

s_grid <- tidyr::expand_grid(
  time_since_death_years = t_seq,
  distance_f = factor(as.character(dist_keep), levels = dist_levels)
) %>%
  mutate(
    distance_m = as.numeric(as.character(distance_f)),  # for labels
    habitat_moist3 = make_col(mf_s, "habitat_moist3", ref_hab_s, n()),
    carcass_size   = make_col(mf_s, "carcass_size",  ref_size_s, n())
  )

s_pred <- pred_ci_cloglog(m_soil, s_grid)

s_pred <- s_pred %>%
  mutate(
    distance_lab = ifelse(distance_m %% 1 == 0,
                          paste0(as.integer(distance_m), "m"),
                          paste0(distance_m, "m")),
    distance_lab = factor(distance_lab,
                          levels = unique(distance_lab[order(distance_m)]))
  )

ggplot(s_pred, aes(x = time_since_death_years, y = p,
                   color = distance_lab, fill = distance_lab, group = distance_lab)) +
  geom_ribbon(aes(ymin = p_lo, ymax = p_hi), alpha = 0.18, color = NA) +
  geom_line(linewidth = 1) +
  facet_wrap(~ distance_lab, nrow = 1) +
  labs(
    x = "Years since death",
    y = "Predicted P(culture+)",
    title = "Predicted Soil Culture Positivity Over Time by Distance from Carcass Center",
    color = "Distance (m)",
    fill  = "Distance (m)"
  ) +
  coord_cartesian(ylim = c(0, 1)) +
  theme(legend.position = "none")

#Nicer Figure Coding#
prep_distance_labels <- function(pred_df, distance_var = "distance_m") {
  stopifnot(distance_var %in% names(pred_df))

  x <- pred_df[[distance_var]]

  x_num <- suppressWarnings(as.numeric(as.character(x)))
  if (all(is.na(x_num))) {
    stop("Couldn't coerce ", distance_var, " to numeric. Check what's in that column.")
  }

  levs <- sort(unique(x_num))
  pred_df %>%
    mutate(
      distance_lab = factor(
        paste0(x_num, "m"),
        levels = paste0(levs, "m")
      )
    )
}

plot_pred_curves_faceted <- function(pred_df, title, subtitle = NULL, caption = NULL, ncol = 3) {
  ggplot(pred_df, aes(x = time_since_death_years, y = p,
                      color = distance_lab, fill = distance_lab)) +
    geom_ribbon(aes(ymin = p_lo, ymax = p_hi), alpha = 0.20, linewidth = 0) +
    geom_line(linewidth = 0.9) +
    facet_wrap(~ distance_lab, ncol = ncol) +
    scale_y_continuous(
      limits = c(0, 1),
      breaks = seq(0, 1, by = 0.25),
      labels = label_number(accuracy = 0.01)
    ) +
    scale_x_continuous(expand = expansion(mult = c(0.01, 0.01))) +
    guides(color = "none", fill = "none") +
    labs(
      x = "Years Since Death",
      y = "Predicted P(Culture+)",
      title = title,
      subtitle = subtitle,
      caption = caption
    ) +
    theme_pub
}
#Nicer Tissue Plot#
p_tiss_pub <- ggplot(t_pred, aes(x = time_since_death_years, y = p)) +
  geom_ribbon(aes(ymin = p_lo, ymax = p_hi), alpha = 0.20, linewidth = 0) +
  geom_line(linewidth = 0.9) +
  scale_y_continuous(
    limits = c(0, 1),
    breaks = seq(0, 1, by = 0.25),
    labels = label_number(accuracy = 0.01)
  ) +
  scale_x_continuous(expand = expansion(mult = c(0.01, 0.01))) +
  labs(
    x = "Years Since Death",
    y = "Predicted P(Culture+)",
    title = "Predicted Tissue Culture Positivity Over Time, Large Carcasses"
  ) +
  theme_pub

p_tiss_pub

#Nicer Soil Plots#
s_pred_pub <- prep_distance_labels(s_pred, distance_var = "distance_m")

p_soil_pub <- plot_pred_curves_faceted(
  s_pred_pub,
  title = "Predicted Soil Culture Positivity Over Time by Distance from Carcass Center",
  subtitle = NULL,
  ncol = 2
)

p_soil_pub
```


# GLMM Models
## Setup & Model Fitting
```{r, GLMM Models, cache=FALSE}

library(glmmTMB)
library(broom.mixed)
library(DHARMa)
library(ggeffects)

#Collapse habitat_class -> 3-level moisture class#
make_habitat_moist3 <- function(habitat_class) {
  case_when(
    habitat_class %in% c("Tundra Barrens") ~ "Dry",
    habitat_class %in% c("Hummock Tundra", "Tundra Meadow") ~ "Medium",
    habitat_class %in% c("Riverbed", "Riparian Tundra", "Tundra Wetland") ~ "Wet",
    TRUE ~ NA_character_
  )
}

#Filter and Recode Data#
dat0 <- dat %>%
  filter(
    carcass_association == "Y",
    substrate %in% c("Tissue", "Soil"),
    !is.na(carcass_id)
  ) %>%
  mutate(
    # outcomes at sample-level (we’ll collapse to visits below)
    cult_pos   = if_else(culture_done == "Y", culture_ac_pos == "Y", NA),  # TRUE/FALSE/NA
    pcr_pos    = if_else(pcr_done == "Y", pcr_ac_pos == "Y", NA),         # TRUE/FALSE/NA
    detect_pos = case_when(
      culture_done == "Y" & culture_ac_pos == "Y" ~ TRUE,                 # culture positive
      culture_done == "Y" & culture_ac_pos != "Y" & pcr_done == "Y" & pcr_ac_pos == "Y" ~ TRUE,  # culture neg + PCR pos
      culture_done == "Y" & culture_ac_pos != "Y" & pcr_done == "Y" & (is.na(pcr_ac_pos) | pcr_ac_pos != "Y") ~ FALSE, # culture neg + PCR done (neg)
      culture_done == "Y" & culture_ac_pos == "Y" & pcr_done != "Y" ~ TRUE, # culture pos, PCR not done
      culture_done == "Y" & culture_ac_pos != "Y" & pcr_done != "Y" ~ NA,   # culture neg, PCR not done => unknown detectability
      culture_done != "Y" & pcr_done == "Y" ~ (pcr_ac_pos == "Y"),          # no culture, PCR only
      TRUE ~ NA
    ),
    habitat_moist3 = make_habitat_moist3(habitat_class),
    habitat_moist3 = factor(habitat_moist3, levels = c("Dry", "Medium", "Wet")),
    carcass_size   = factor(carcass_size, levels = c(1, 2), labels = c("Small", "Large")),
    year_of_death  = as.numeric(year_of_death),
    sample_year    = as.numeric(sample_year)
  )

#Collapse to "Visits"#
tiss_vis <- dat0 %>%
  filter(substrate == "Tissue") %>%
  group_by(carcass_id, time_since_death_years) %>%
  summarise(
    y_cult   = if_else(any(culture_done == "Y"), as.integer(any(culture_ac_pos == "Y", na.rm = TRUE)), NA_integer_),
    y_detect = if_else(any(culture_done == "Y" | pcr_done == "Y"), as.integer(any(detect_pos == TRUE, na.rm = TRUE)), NA_integer_),
    carcass_size   = first(na.omit(carcass_size)),
    habitat_moist3 = first(na.omit(habitat_moist3)),
    year_of_death  = first(na.omit(year_of_death)),
    sample_year    = first(na.omit(sample_year)),
    .groups = "drop"
  ) %>%
  mutate(
    time_c = time_since_death_years - mean(time_since_death_years, na.rm = TRUE),
    year_c = year_of_death - mean(year_of_death, na.rm = TRUE)
  )

soil_vis <- dat0 %>%
  filter(substrate == "Soil") %>%
  group_by(carcass_id, distance_m, time_since_death_years) %>%
  summarise(
    y_cult   = if_else(any(culture_done == "Y"), as.integer(any(culture_ac_pos == "Y", na.rm = TRUE)), NA_integer_),
    y_detect = if_else(any(culture_done == "Y" | pcr_done == "Y"), as.integer(any(detect_pos == TRUE, na.rm = TRUE)), NA_integer_),
    carcass_size   = first(na.omit(carcass_size)),
    habitat_moist3 = first(na.omit(habitat_moist3)),
    year_of_death  = first(na.omit(year_of_death)),
    sample_year    = first(na.omit(sample_year)),
    .groups = "drop"
  ) %>%
  mutate(
    time_c = time_since_death_years - mean(time_since_death_years, na.rm = TRUE),
    dist_c = distance_m - mean(distance_m, na.rm = TRUE),
    year_c = year_of_death - mean(year_of_death, na.rm = TRUE),
    distance_f = factor(distance_m, levels = c(0,1,3,5,15))
  )

link_fun <- "logit"

#Subset Soil Distances to 0m-5m to avoid model separation and avoid confusion with uphill/downhill gradient#
soil_vis2 <- soil_vis %>%
  mutate(
    distance_f = factor(distance_m, levels = c(0, 1, 3, 5))
  )

#####################
#####FULL MODELS#####
#####################

###TISSUE VIABILITY MODEL###
m_tiss_cult <- glmmTMB(
  y_cult ~ time_c + carcass_size + habitat_moist3 + (1 | carcass_id),
  family = binomial(link = link_fun),
  data   = tiss_vis %>% filter(!is.na(y_cult), !is.na(time_c), !is.na(carcass_size), !is.na(habitat_moist3))
)

###TISSUE DETECTABILITY MODEL###
m_tiss_det <- glmmTMB(
  y_detect ~ time_c + carcass_size + habitat_moist3 + (1 | carcass_id),
  family = binomial(link = link_fun),
  data   = tiss_vis %>% filter(!is.na(y_detect), !is.na(time_c), !is.na(carcass_size), !is.na(habitat_moist3))
)

###SOIL VIABILITY MODELS###
m_soil_cult <- glmmTMB(
  y_cult ~ time_c + distance_f + carcass_size + habitat_moist3 + (1 | carcass_id),
  family = binomial(link = link_fun),
  data   = soil_vis %>% filter(!is.na(y_cult), !is.na(time_c), !is.na(distance_f), !is.na(carcass_size), !is.na(habitat_moist3))
)

#with time*dist interaction#
m_soil_cult_timedist <- update(m_soil_cult, . ~ . + time_c:distance_f)

#distances limited to 0-5m#
m_soil_cult_df <- glmmTMB(
  y_cult ~ time_c + distance_f + carcass_size + habitat_moist3 + (1 | carcass_id),
  family = binomial(link = link_fun),
  data = soil_vis2 %>%
    filter(!is.na(y_cult), !is.na(time_c), !is.na(distance_f),
           !is.na(carcass_size), !is.na(habitat_moist3))
)

#distance limited with time*dist interaction#
m_soil_cult_df_timedist <- update(m_soil_cult_df, . ~ . + time_c:distance_f)

###SOIL DETECTABILITY MODELS###
m_soil_det <- glmmTMB(
  y_detect ~ time_c + distance_f + carcass_size + habitat_moist3 + (1 | carcass_id),
  family = binomial(link = link_fun),
  data   = soil_vis %>% filter(!is.na(y_detect), !is.na(time_c), !is.na(distance_f), !is.na(carcass_size), !is.na(habitat_moist3))
)

#with time*dist interaction#
m_soil_det_timedist <- update(m_soil_det, . ~ . + time_c:distance_f)

#distances limited to 0-5m#
m_soil_det_df <- glmmTMB(
  y_detect ~ time_c + distance_f + carcass_size + habitat_moist3 + (1 | carcass_id),
  family = binomial(link = link_fun),
  data = soil_vis2 %>%
    filter(!is.na(y_detect), !is.na(time_c), !is.na(distance_f),
           !is.na(carcass_size), !is.na(habitat_moist3))
)

#distance limited with time*dist interaction#
m_soil_det_df_timedist <- update(m_soil_det_df, . ~ . + time_c:distance_f)

print(m_tiss_cult)
print(m_tiss_det)
print(m_soil_cult)
print(m_soil_cult_timedist)
print(m_soil_cult_df)
print(m_soil_cult_df_timedist)
print(m_soil_det)
print(m_soil_det_timedist)
print(m_soil_det_df)
print(m_soil_det_df_timedist)

###MODEL DIAGNOSTICS###
sim_tiss_cult <- simulateResiduals(m_tiss_cult); plot(sim_tiss_cult)
sim_tiss_det  <- simulateResiduals(m_tiss_det);  plot(sim_tiss_det)
sim_soil_cult <- simulateResiduals(m_soil_cult); plot(sim_soil_cult)
sim_soil_cult_timedist <- simulateResiduals(m_soil_cult_timedist); plot(sim_soil_cult_timedist)
sim_soil_cult_df <- simulateResiduals(m_soil_cult_df); plot(sim_soil_cult_df)
sim_soil_cult_df_timedist <- simulateResiduals(m_soil_cult_df_timedist); plot(sim_soil_cult_df_timedist)
sim_soil_det  <- simulateResiduals(m_soil_det);  plot(sim_soil_det)
sim_soil_det_timedist <- simulateResiduals(m_soil_det_timedist);  plot(sim_soil_det_timedist)
sim_soil_det_df <- simulateResiduals(m_soil_det_df); plot(sim_soil_det_df)
sim_soil_det_df_timedist <- simulateResiduals(m_soil_det_df_timedist); plot(sim_soil_det_df_timedist)

###MODEL TABLES###
tidy_or <- function(m, model_name) {
  broom.mixed::tidy(m, effects = "fixed", conf.int = TRUE, conf.method = "Wald") %>%
    mutate(
      model = model_name,
      OR    = exp(estimate),
      OR_lo = exp(conf.low),
      OR_hi = exp(conf.high)
    ) %>%
    dplyr::select(model, term, estimate, std.error, statistic, p.value, OR, OR_lo, OR_hi)
}

tab_models <- bind_rows(
  tidy_or(m_tiss_cult, "Tissue | Viability (culture+)"),
  tidy_or(m_tiss_det,  "Tissue | Detectability (culture or PCR+)"),
  tidy_or(m_soil_cult, "Soil   | Viability (culture+)"),
  tidy_or(m_soil_cult_timedist, "Soil   | Viability (culture+) w/ Time*Dist Interaction"),
  tidy_or(m_soil_cult_df, "Soil   | Viability (culture+) w/ Limited Distances"),
  tidy_or(m_soil_cult_df_timedist, "Soil   | Viability (culture+) w/ Limited Distances & Time*Dist Interaction"),
  tidy_or(m_soil_det,  "Soil   | Detectability (culture or PCR+)"),
  tidy_or(m_soil_det_timedist,  "Soil   | Detectability (culture or PCR+) w/Time*Dist Interaction"),
  tidy_or(m_soil_det_df, "Soil   | Detectability (culture or PCR+) w/ Limited Distances"),
  tidy_or(m_soil_det_df_timedist, "Soil   | Detectability (culture or PCR+) w/ Limited Distances & Time*Dist Interaction")
)

tab_models %>% print(n = Inf)

tab_models_int <- bind_rows(
  tidy_or(m_soil_cult_df_timedist, "Soil   | Viability (culture+) w/ Limited Distances & Time*Dist Interaction"),
  tidy_or(m_soil_det_df_timedist, "Soil   | Detectability (culture or PCR+) w/ Limited Distances & Time*Dist Interaction")
)

tab_models_int %>% print(n = Inf)

tab_models_chosen <- bind_rows(
  tidy_or(m_tiss_cult, "Tissue | Viability (culture+)"),
  tidy_or(m_tiss_det,  "Tissue | Detectability (culture or PCR+)"),
  tidy_or(m_soil_cult_df, "Soil   | Viability (culture+) w/ Limited Distances"),
  tidy_or(m_soil_det_df, "Soil   | Detectability (culture or PCR+) w/ Limited Distances"))

tab_models_chosen %>% print(n = Inf)

##########################
##### NO-TIME MODELS #####
##########################

#Build no-time visit tables#
tiss_vis_nt <- dat0 %>%
  filter(substrate == "Tissue") %>%
  group_by(carcass_id, sample_year) %>%
  summarise(
    y_cult   = if_else(any(culture_done == "Y"),
                       as.integer(any(culture_ac_pos == "Y", na.rm = TRUE)),
                       NA_integer_),
    y_detect = if_else(any(culture_done == "Y" | pcr_done == "Y"),
                       as.integer(any(detect_pos == TRUE, na.rm = TRUE)),
                       NA_integer_),
    carcass_size   = first(na.omit(carcass_size)),
    habitat_moist3 = first(na.omit(habitat_moist3)),
    .groups = "drop"
  )

soil_vis_nt <- dat0 %>%
  filter(substrate == "Soil") %>%
  group_by(carcass_id, sample_year, distance_m) %>%
  summarise(
    y_cult   = if_else(any(culture_done == "Y"),
                       as.integer(any(culture_ac_pos == "Y", na.rm = TRUE)),
                       NA_integer_),
    y_detect = if_else(any(culture_done == "Y" | pcr_done == "Y"),
                       as.integer(any(detect_pos == TRUE, na.rm = TRUE)),
                       NA_integer_),
    carcass_size   = first(na.omit(carcass_size)),
    habitat_moist3 = first(na.omit(habitat_moist3)),
    .groups = "drop"
  ) %>%
  mutate(
    distance_f = factor(distance_m, levels = c(0, 1, 3, 5, 15))
  )

soil_vis_nt_df <- soil_vis_nt %>%
  filter(distance_m %in% c(0, 1, 3, 5)) %>%
  mutate(distance_f = factor(distance_m, levels = c(0, 1, 3, 5)))

### TISSUE VIABILITY (NO TIME) ###
m_tiss_cult_nt <- glmmTMB(
  y_cult ~ carcass_size + habitat_moist3 + (1 | carcass_id),
  family = binomial(link = link_fun),
  data = tiss_vis_nt %>%
    filter(!is.na(y_cult),
           !is.na(carcass_size),
           !is.na(habitat_moist3))
)

### TISSUE DETECTABILITY (NO TIME) ###
m_tiss_det_nt <- glmmTMB(
  y_detect ~ carcass_size + habitat_moist3 + (1 | carcass_id),
  family = binomial(link = link_fun),
  data = tiss_vis_nt %>%
    filter(!is.na(y_detect),
           !is.na(carcass_size),
           !is.na(habitat_moist3))
)

### SOIL VIABILITY (NO TIME) ###
m_soil_cult_nt_df <- glmmTMB(
  y_cult ~ distance_f + carcass_size + habitat_moist3 + (1 | carcass_id),
  family = binomial(link = link_fun),
  data = soil_vis_nt_df %>%
    filter(!is.na(y_cult),
           !is.na(distance_f),
           !is.na(carcass_size),
           !is.na(habitat_moist3))
)

### SOIL DETECTABILITY (NO TIME) ###
m_soil_det_nt_df <- glmmTMB(
  y_detect ~ distance_f + carcass_size + habitat_moist3 + (1 | carcass_id),
  family = binomial(link = link_fun),
  data = soil_vis_nt_df %>%
    filter(!is.na(y_detect),
           !is.na(distance_f),
           !is.na(carcass_size),
           !is.na(habitat_moist3))
)


print(m_tiss_cult_nt)
print(m_tiss_det_nt)
print(m_soil_cult_nt_df)
print(m_soil_det_nt_df)

#NO TIME CARCASS SITES AND VISITS USED#
model_counts <- bind_rows(
  tiss_vis_nt %>%
    filter(!is.na(y_cult), !is.na(carcass_size), !is.na(habitat_moist3)) %>%
    summarise(model = "Tissue viability (no time)",
              visits = n(),
              carcass_sites = n_distinct(carcass_id)),
  tiss_vis_nt %>%
    filter(!is.na(y_detect), !is.na(carcass_size), !is.na(habitat_moist3)) %>%
    summarise(model = "Tissue detectability (no time)",
              visits = n(),
              carcass_sites = n_distinct(carcass_id)),
  soil_vis_nt_df %>%
    filter(!is.na(y_cult), !is.na(distance_f), !is.na(carcass_size), !is.na(habitat_moist3)) %>%
    summarise(model = "Soil viability 0–5 m (no time)",
              visits = n(),
              carcass_sites = n_distinct(carcass_id)),
  soil_vis_nt_df %>%
    filter(!is.na(y_detect), !is.na(distance_f), !is.na(carcass_size), !is.na(habitat_moist3)) %>%
    summarise(model = "Soil detectability 0–5 m (no time)",
              visits = n(),
              carcass_sites = n_distinct(carcass_id))
)

model_counts

#NO TIME MODEL TABLES#

tidy_or <- function(m, model_name) {
  broom.mixed::tidy(m, effects = "fixed", conf.int = TRUE, conf.method = "Wald") %>%
    mutate(
      model = model_name,
      OR    = exp(estimate),
      OR_lo = exp(conf.low),
      OR_hi = exp(conf.high)
    ) %>%
    dplyr::select(model, term, estimate, std.error, statistic, p.value, OR, OR_lo, OR_hi)
}

tab_models_nt <- bind_rows(
  tidy_or(m_tiss_cult_nt,    "Tissue viability (no time)"),
  tidy_or(m_tiss_det_nt,     "Tissue detectability (no time)"),
  tidy_or(m_soil_cult_nt_df, "Soil viability 0–5 m (no time)"),
  tidy_or(m_soil_det_nt_df,  "Soil detectability 0–5 m (no time)")
)

print(tab_models_nt, n=Inf)
```

## Soil Predicted-Curve Plots
```{r, GLMM Models: Soil Predicted-Curve Plots, cache=FALSE}

library(patchwork)

#Build plotting dataset#
soil_plot_dat <- soil_vis2 %>%
  filter(
    distance_m %in% c(0, 1, 3, 5),
    !is.na(time_since_death_years),
    !is.na(time_c),
    !is.na(distance_f),
    !is.na(carcass_size),
    !is.na(habitat_moist3)
  )

#Pull most common fixed-effect levels#
mf_s_glmm <- model.frame(m_soil_cult_df)

ref_hab_s_glmm  <- names(sort(table(mf_s_glmm$habitat_moist3), decreasing = TRUE))[1]
ref_size_s_glmm <- names(sort(table(mf_s_glmm$carcass_size),  decreasing = TRUE))[1]

#Time sequence from plotting data#
t_seq_glmm <- seq(
  min(soil_plot_dat$time_since_death_years, na.rm = TRUE),
  max(soil_plot_dat$time_since_death_years, na.rm = TRUE),
  length.out = 200
)

t_mean_glmm <- mean(soil_plot_dat$time_since_death_years, na.rm = TRUE)

#Prediction grid#
soil_pred_grid <- tidyr::expand_grid(
  time_since_death_years = t_seq_glmm,
  distance_f = factor(c(0, 1, 3, 5), levels = c(0, 1, 3, 5))
) %>%
  mutate(
    time_c         = time_since_death_years - t_mean_glmm,
    habitat_moist3 = factor(ref_hab_s_glmm,  levels = levels(mf_s_glmm$habitat_moist3)),
    carcass_size   = factor(ref_size_s_glmm, levels = levels(mf_s_glmm$carcass_size))
  )

#Get population-level predicted probabilities + 95% CI#
predict_glmm_curve <- function(model, newdata, outcome_label) {
  pr <- predict(
    model,
    newdata = newdata,
    type    = "link",
    se.fit  = TRUE,
    re.form = NA
  )

  out <- newdata %>%
    mutate(
      eta     = as.numeric(pr$fit),
      se      = as.numeric(pr$se.fit),
      p       = plogis(eta),
      p_lo    = plogis(eta - 1.96 * se),
      p_hi    = plogis(eta + 1.96 * se),
      outcome = outcome_label
    )

  prep_distance_labels(out, distance_var = "distance_f")
}

#Predictions#
soil_pred_viab <- predict_glmm_curve(
  m_soil_cult_df,
  soil_pred_grid,
  "Viability"
)

soil_pred_det <- predict_glmm_curve(
  m_soil_det_df,
  soil_pred_grid,
  "Detectability"
)

#Panel plotting#
plot_glmm_soil_panel <- function(pred_df, ylab_text, panel_title = NULL) {
  ggplot(
    pred_df,
    aes(
      x = time_since_death_years,
      y = p,
      color = distance_lab,
      fill  = distance_lab
    )
  ) +
    geom_ribbon(aes(ymin = p_lo, ymax = p_hi), alpha = 0.18, linewidth = 0) +
    geom_line(linewidth = 0.9) +
    facet_wrap(~ distance_lab, nrow = 1) +
    scale_y_continuous(
      limits = c(0, 1),
      breaks = seq(0, 1, by = 0.25),
      labels = label_number(accuracy = 0.01)
    ) +
    scale_x_continuous(
      expand = expansion(mult = c(0.01, 0.01))
    ) +
    guides(color = "none", fill = "none") +
    labs(
      x = "Years Since Death",
      y = ylab_text,
      title = panel_title
    ) +
    theme_pub
}

p_soil_glmm_viab <- plot_glmm_soil_panel(
  soil_pred_viab,
  ylab_text  = "Predicted P(Culture+)",
  panel_title = "Predicted Soil Viability Over Time"
)

p_soil_glmm_viab

p_soil_glmm_det <- plot_glmm_soil_panel(
  soil_pred_det,
  ylab_text  = "Predicted P(Culture or PCR+)",
  panel_title = "Predicted Soil Detectability Over Time"
)

p_soil_glmm_det

p_soil_glmm_two_panel <- patchwork::wrap_plots(
  p_soil_glmm_viab,
  p_soil_glmm_det,
  ncol = 1
)

p_soil_glmm_two_panel
```

##Tissue Predicted-Curve Plots
```{r, GLMM Models: Tissue Predicted-Curve Plots, cache=FALSE}

library(patchwork)

# Build plotting dataset using the same variables included in the tissue GLMMs
tiss_plot_dat <- tiss_vis %>%
  filter(
    !is.na(time_since_death_years),
    !is.na(time_c),
    !is.na(carcass_size),
    !is.na(habitat_moist3)
  )

# Pull most common fixed-effect levels from the model frame
mf_t_glmm <- model.frame(m_tiss_cult)

ref_hab_t_glmm <- names(sort(table(mf_t_glmm$habitat_moist3), decreasing = TRUE))[1]
ref_size_t_glmm <- names(sort(table(mf_t_glmm$carcass_size), decreasing = TRUE))[1]

# Time sequence from tissue plotting data
t_seq_tiss_glmm <- seq(
  min(tiss_plot_dat$time_since_death_years, na.rm = TRUE),
  max(tiss_plot_dat$time_since_death_years, na.rm = TRUE),
  length.out = 200
)

t_mean_tiss_glmm <- mean(tiss_plot_dat$time_since_death_years, na.rm = TRUE)

# Prediction grid at most common reference categories
tiss_pred_grid <- tibble::tibble(
  time_since_death_years = t_seq_tiss_glmm,
  time_c = time_since_death_years - t_mean_tiss_glmm,
  habitat_moist3 = factor(
    ref_hab_t_glmm,
    levels = levels(mf_t_glmm$habitat_moist3)
  ),
  carcass_size = factor(
    ref_size_t_glmm,
    levels = levels(mf_t_glmm$carcass_size)
  )
)

# General GLMM prediction helper for tissue curves
predict_glmm_tissue_curve <- function(model, newdata, outcome_label) {
  pr <- predict(
    model,
    newdata = newdata,
    type    = "link",
    se.fit  = TRUE,
    re.form = NA
  )

  newdata %>%
    mutate(
      eta     = as.numeric(pr$fit),
      se      = as.numeric(pr$se.fit),
      p       = plogis(eta),
      p_lo    = plogis(eta - 1.96 * se),
      p_hi    = plogis(eta + 1.96 * se),
      outcome = outcome_label
    )
}

# Predictions
tiss_pred_viab <- predict_glmm_tissue_curve(
  m_tiss_cult,
  tiss_pred_grid,
  "Viability"
)

tiss_pred_det <- predict_glmm_tissue_curve(
  m_tiss_det,
  tiss_pred_grid,
  "Detectability"
)

# Panel plotting function
plot_glmm_tissue_panel <- function(pred_df, ylab_text, panel_title = NULL) {
  ggplot(pred_df, aes(x = time_since_death_years, y = p)) +
    geom_ribbon(aes(ymin = p_lo, ymax = p_hi), alpha = 0.18, linewidth = 0) +
    geom_line(linewidth = 0.9) +
    scale_y_continuous(
      limits = c(0, 1),
      breaks = seq(0, 1, by = 0.25),
      labels = scales::label_number(accuracy = 0.01)
    ) +
    scale_x_continuous(
      expand = expansion(mult = c(0.01, 0.01))
    ) +
    labs(
      x = "Years Since Death",
      y = ylab_text,
      title = panel_title
    ) +
    theme_pub
}

# Individual panels
p_tiss_glmm_viab <- plot_glmm_tissue_panel(
  tiss_pred_viab,
  ylab_text = "Predicted P(Culture+)",
  panel_title = "Predicted Tissue Viability Over Time"
)

p_tiss_glmm_det <- plot_glmm_tissue_panel(
  tiss_pred_det,
  ylab_text = "Predicted P(Culture or PCR+)",
  panel_title = "Predicted Tissue Detectability Over Time"
)

# Side-by-side figure
p_tiss_glmm_two_panel <- patchwork::wrap_plots(
  p_tiss_glmm_viab,
  p_tiss_glmm_det,
  ncol = 2
)

p_tiss_glmm_two_panel

ggsave(
  filename = "Tissue Predicted Curves.png",
  plot = p_tiss_glmm_two_panel,
  width = 10,
  height = 5,
  dpi = 600
)

```
# Soil Chemistry Analyses
## Data Setup
```{r, Soil Chemistry: Data Setup, cache=FALSE}
library(tidyr)
library(vegan)
library(tibble)
library(purrr)
library(glmmTMB)
library(broom.mixed)
library(ggplot2)

# Habitat Moisture Classes 
make_habitat_moist3 <- function(x) {
  x <- as.character(x)
  out <- dplyr::case_when(
    x == "Tundra Barrens" ~ "Dry",
    x %in% c("Hummock Tundra", "Tundra Meadow") ~ "Medium",
    x %in% c("Riparian Tundra", "Riverbed", "Tundra Wetland") ~ "Wet",
    TRUE ~ NA_character_
  )

  factor(out, levels = c("Dry", "Medium", "Wet"))
}

# Soil chemistry variables
chem_vars_raw <- c(
  "soil_moisture",
  "soil_water_hold_cap",
  "soil_pH",
  "soil_soluble_C",
  "soil_soluble_orgC",
  "soil_soluble_inorgC",
  "soil_N",
  "soil_CN_ratio",
  "soil_Cl",
  "soil_K",
  "soil_P"
)

# Variables to log-adjust prior to multivariate / delta work 
log_vars <- c(
  "soil_water_hold_cap",
  "soil_soluble_C",
  "soil_soluble_orgC",
  "soil_soluble_inorgC",
  "soil_N",
  "soil_CN_ratio",
  "soil_Cl",
  "soil_K",
  "soil_P"
)

# Build raw paired soil chemistry dataset
soil_chem <- dat %>%
  filter(
    carcass_association == "Y",
    substrate == "Soil",
    sample_year %in% c(2023, 2024),
    distance_m %in% c(0, 15)
  ) %>%
  filter(if_any(all_of(chem_vars_raw), ~ !is.na(.x))) %>%
  mutate(
    sample_year = as.integer(sample_year),
    distance_m = as.integer(distance_m),
    carcass_id = as.character(carcass_id),
    sample_id = as.character(sample_id),
    habitat_class = as.character(habitat_class),
    habitat_moist3 = make_habitat_moist3(habitat_class),
    distance_lab = factor(ifelse(distance_m == 0, "0m", "15m"),
                          levels = c("0m", "15m"))
  )

# Apply log1p transformations
soil_chem_tr <- soil_chem %>%
  mutate(
    across(all_of(log_vars), log1p, .names = "{.col}_log")
  )

# Variables used in multivariate analyses
chem_vars_use <- c(
  "soil_moisture",
  "soil_water_hold_cap_log",
  "soil_pH",
  "soil_soluble_C_log",
  "soil_soluble_orgC_log",
  "soil_soluble_inorgC_log",
  "soil_N_log",
  "soil_CN_ratio_log",
  "soil_Cl_log",
  "soil_K_log",
  "soil_P_log"
)

# Standardize transformed raw paired dataset
soil_chem_std <- soil_chem_tr %>%
  drop_na(all_of(chem_vars_use)) %>%
  mutate(
    across(
      all_of(chem_vars_use),
      ~ as.numeric(scale(.x)),
      .names = "{.col}_z"
    )
  )

chem_vars_z <- paste0(chem_vars_use, "_z")

# Pairwise PERMANOVA helper function
pairwise_permanova <- function(df, group_var, vars_z, strata_var = NULL,
                               nperm = 9999) {

  groups <- levels(droplevels(factor(df[[group_var]])))
  comps  <- combn(groups, 2, simplify = FALSE)

  out <- map_dfr(comps, function(cc) {

    dsub <- df %>%
      filter(.data[[group_var]] %in% cc) %>%
      mutate(.grp = droplevels(factor(.data[[group_var]])))

    X <- as.matrix(dsub[, vars_z])

    fit <- adonis2(
      X ~ .grp,
      data = dsub,
      method = "euclidean",
      permutations = nperm,
      strata = if (!is.null(strata_var)) dsub[[strata_var]] else NULL
    )

    tibble(
      group1  = cc[1],
      group2  = cc[2],
      n       = nrow(dsub),
      F_model = fit$F[1],
      R2      = fit$R2[1],
      p       = fit$`Pr(>F)`[1]
    )
  })

  out %>%
    mutate(p_adj_BH = p.adjust(p, method = "BH")) %>%
    arrange(p_adj_BH, p)
}

# Global PERMANOVA + dispersion wrapper
run_perm_suite <- function(df, group_var, vars_z, strata_var = NULL,
                           nperm = 9999) {

  df <- df %>%
    filter(!is.na(.data[[group_var]])) %>%
    mutate(.grp = droplevels(factor(.data[[group_var]])))

  X <- as.matrix(df[, vars_z])
  D <- vegdist(X, method = "euclidean")

  perm <- adonis2(
    X ~ .grp,
    data = df,
    method = "euclidean",
    permutations = nperm,
    strata = if (!is.null(strata_var)) df[[strata_var]] else NULL
  )

  bd <- betadisper(D, group = df$.grp, type = "centroid")
  bd_perm <- permutest(bd, permutations = nperm)

  pw_perm <- pairwise_permanova(
    df = df,
    group_var = group_var,
    vars_z = vars_z,
    strata_var = strata_var,
    nperm = nperm
  )

  list(
    permanova = perm,
    permdisp = bd_perm,
    betadisper = bd,
    pairwise_permanova = pw_perm
  )
}

# Build paired dataset for deltas

soil_pairs <- soil_chem_tr %>%
  dplyr::select(
    carcass_id, sample_year, time_since_death_years,
    habitat_moist3, distance_lab, all_of(chem_vars_use)
  ) %>%
  pivot_wider(
    names_from  = distance_lab,
    values_from = all_of(chem_vars_use),
    names_sep   = "_"
  ) %>%
  mutate(
    years_since_death_f = case_when(
      time_since_death_years == 2 ~ "2y",
      time_since_death_years == 3 ~ "3y",
      TRUE ~ NA_character_
    ),
    years_since_death_f = factor(years_since_death_f, levels = c("2y", "3y"))
  )

# Absolute delta = 0m - 15m
for (v in chem_vars_use) {
  soil_pairs[[paste0("delta_", v)]] <-
    soil_pairs[[paste0(v, "_0m")]] - soil_pairs[[paste0(v, "_15m")]]
}

# Percentage delta = ((0m - 15m) / 15m) * 100
for (v in chem_vars_use) {
  denom <- soil_pairs[[paste0(v, "_15m")]]

  soil_pairs[[paste0("pct_delta_", v)]] <-
    ifelse(
      !is.na(denom) & denom != 0,
      ((soil_pairs[[paste0(v, "_0m")]] - denom) / denom) * 100,
      NA_real_
    )
}

delta_vars_abs <- paste0("delta_", chem_vars_use)
delta_vars_pct <- paste0("pct_delta_", chem_vars_use)

soil_delta <- soil_pairs

soil_delta_tsd <- soil_delta %>%
  filter(years_since_death_f %in% c("2y", "3y"))

# Quick checks
soil_chem %>% count(sample_year, distance_lab)
soil_chem_std %>% count(sample_year, distance_lab, habitat_moist3)
soil_delta_tsd %>% count(years_since_death_f, habitat_moist3)


```

##Soil Chemistry Sample Summary Data Table##
```{r, Soil Chemistry: Supplementary Table, cache=FALSE}

library(tidyverse)
library(stringr)
library(readr)

#------------------------------------------------------------
# Formatting helpers
#------------------------------------------------------------

fmt_n_pct <- function(n, d, digits = 1) {
  ifelse(
    is.na(d) | d == 0,
    "",
    sprintf(paste0("%d (%.", digits, "f)"), n, 100 * n / d)
  )
}

fmt_pos_tested <- function(pos, tested, digits = 1) {
  ifelse(
    is.na(tested) | tested == 0,
    "",
    sprintf(paste0("%d/%d (%.", digits, "f)"), pos, tested, 100 * pos / tested)
  )
}

fmt_n_units <- function(n, units) {
  ifelse(
    is.na(n),
    "",
    sprintf("%d (%d)", n, units)
  )
}

clean_yn <- function(x) {
  str_to_upper(str_trim(as.character(x)))
}

#------------------------------------------------------------
# Build Supplementary Table 7 dataset from soil_chem ONLY
#------------------------------------------------------------

supp7_dat <- soil_chem %>%
  mutate(
    unique_id = as.character(unique_id),
    site_id = as.character(site_id),
    carcass_id = as.character(carcass_id),
    
    sample_year = as.integer(sample_year),
    distance_m = as.integer(distance_m),
    
    year_of_death = suppressWarnings(as.numeric(year_of_death)),
    time_since_death_years = suppressWarnings(as.numeric(time_since_death_years)),
    
    culture_done_chr = clean_yn(culture_done),
    culture_ac_pos_chr = clean_yn(culture_ac_pos),
    pcr_done_chr = clean_yn(pcr_done),
    pcr_ac_pos_chr = clean_yn(pcr_ac_pos),
    ac_any_pos_chr = clean_yn(ac_any_pos),
    
    distance_lab = case_when(
      distance_m == 0 ~ "0 m",
      distance_m == 15 ~ "15 m",
      TRUE ~ paste0(distance_m, " m")
    ),
    
    culture_tested = culture_done_chr == "Y" | culture_ac_pos_chr %in% c("Y", "N"),
    culture_pos = culture_ac_pos_chr == "Y",
    
    pcr_tested = pcr_done_chr == "Y" | pcr_ac_pos_chr %in% c("Y", "N"),
    pcr_pos = pcr_ac_pos_chr == "Y",
    
    any_tested = culture_tested | pcr_tested,
    any_pos = case_when(
      ac_any_pos_chr %in% c("Y", "N") ~ ac_any_pos_chr == "Y",
      TRUE ~ culture_pos | pcr_pos
    ),
    
    time_known = !is.na(year_of_death) & !is.na(time_since_death_years)
  ) %>%
  filter(
    sample_year %in% c(2023, 2024),
    distance_m %in% c(0, 15),
    !is.na(carcass_id),
    
    # Critical update:
    # remove carcasses without known death timing
    time_known,
    
    # Align with the inferential soil chemistry analyses:
    # 2023 = 2 years since death, 2024 = 3 years since death
    time_since_death_years %in% c(2, 3)
  )

#------------------------------------------------------------
# Diagnostics: confirm what is included/excluded
#------------------------------------------------------------

supp7_removed_unknown_death <- soil_chem %>%
  mutate(
    sample_year = as.integer(sample_year),
    distance_m = as.integer(distance_m),
    year_of_death = suppressWarnings(as.numeric(year_of_death)),
    time_since_death_years = suppressWarnings(as.numeric(time_since_death_years))
  ) %>%
  filter(
    sample_year %in% c(2023, 2024),
    distance_m %in% c(0, 15),
    is.na(year_of_death) | is.na(time_since_death_years)
  ) %>%
  distinct(
    carcass_id,
    site_id,
    sample_year,
    distance_m,
    year_of_death,
    time_since_death_years
  ) %>%
  arrange(carcass_id, sample_year, distance_m)

supp7_removed_unknown_death

supp7_diagnostic <- supp7_dat %>%
  group_by(distance_lab, sample_year) %>%
  summarise(
    samples_n = n(),
    unique_carcasses_n = n_distinct(carcass_id),
    known_time_of_death_n = sum(time_known, na.rm = TRUE),
    
    culture_pos_n = sum(culture_pos, na.rm = TRUE),
    culture_tested_n = sum(culture_tested, na.rm = TRUE),
    
    pcr_pos_n = sum(pcr_pos, na.rm = TRUE),
    pcr_tested_n = sum(pcr_tested, na.rm = TRUE),
    
    any_pos_n = sum(any_pos, na.rm = TRUE),
    any_tested_n = sum(any_tested, na.rm = TRUE),
    
    .groups = "drop"
  ) %>%
  arrange(
    factor(distance_lab, levels = c("0 m", "15 m")),
    sample_year
  )

supp7_diagnostic

#------------------------------------------------------------
# Summaries by distance-year
#------------------------------------------------------------

supp7_years <- supp7_dat %>%
  group_by(distance_lab, sample_year) %>%
  summarise(
    samples_n = n(),
    unique_carcasses_n = n_distinct(carcass_id),
    time_known_n = sum(time_known, na.rm = TRUE),
    
    culture_pos_n = sum(culture_pos, na.rm = TRUE),
    culture_tested_n = sum(culture_tested, na.rm = TRUE),
    
    pcr_pos_n = sum(pcr_pos, na.rm = TRUE),
    pcr_tested_n = sum(pcr_tested, na.rm = TRUE),
    
    any_pos_n = sum(any_pos, na.rm = TRUE),
    any_tested_n = sum(any_tested, na.rm = TRUE),
    
    .groups = "drop"
  ) %>%
  mutate(sample_year = as.character(sample_year))

supp7_totals <- supp7_dat %>%
  group_by(distance_lab) %>%
  summarise(
    sample_year = "TOTAL",
    samples_n = n(),
    unique_carcasses_n = n_distinct(carcass_id),
    time_known_n = sum(time_known, na.rm = TRUE),
    
    culture_pos_n = sum(culture_pos, na.rm = TRUE),
    culture_tested_n = sum(culture_tested, na.rm = TRUE),
    
    pcr_pos_n = sum(pcr_pos, na.rm = TRUE),
    pcr_tested_n = sum(pcr_tested, na.rm = TRUE),
    
    any_pos_n = sum(any_pos, na.rm = TRUE),
    any_tested_n = sum(any_tested, na.rm = TRUE),
    
    .groups = "drop"
  )

supp7_stats <- bind_rows(
  supp7_years,
  supp7_totals
)

#------------------------------------------------------------
# Table shell matching Supplementary Table 7 skeleton
#------------------------------------------------------------

supp7_shell <- tribble(
  ~distance_lab, ~sample_year,
  "0 m",  "2023",
  "0 m",  "2024",
  "0 m",  "TOTAL",
  "15 m", "2023",
  "15 m", "2024",
  "15 m", "TOTAL"
)

supp_table_7 <- supp7_shell %>%
  left_join(
    supp7_stats,
    by = c("distance_lab", "sample_year")
  ) %>%
  mutate(
    across(
      c(
        samples_n,
        unique_carcasses_n,
        time_known_n,
        culture_pos_n,
        culture_tested_n,
        pcr_pos_n,
        pcr_tested_n,
        any_pos_n,
        any_tested_n
      ),
      ~ replace_na(.x, 0)
    ),
    
    `Sampling Distance` = distance_lab,
    Year = sample_year,
    `Samples, n (Unique Carcasses*)` = fmt_n_units(samples_n, unique_carcasses_n),
    `Known Time of Death*, n (%)` = fmt_n_pct(time_known_n, samples_n),
    `Culture + / n Tested (%)` = fmt_pos_tested(culture_pos_n, culture_tested_n),
    `PCR + / n Tested (%)` = fmt_pos_tested(pcr_pos_n, pcr_tested_n),
    `Any Test + / n Tested (%)` = fmt_pos_tested(any_pos_n, any_tested_n)
  ) %>%
  select(
    `Sampling Distance`,
    Year,
    `Samples, n (Unique Carcasses*)`,
    `Known Time of Death*, n (%)`,
    `Culture + / n Tested (%)`,
    `PCR + / n Tested (%)`,
    `Any Test + / n Tested (%)`
  )

supp_table_7

#------------------------------------------------------------
# Display
#------------------------------------------------------------

knitr::kable(
  supp_table_7,
  caption = "Supplementary Table 7. Summary of soil samples assessed for edaphic factors stratified by sampling distance."
)
```

##Model Covariate Summary Table##
```{r, Covariate Summary Table, cache=FALSE}

library(tidyverse)
library(stringr)
library(readr)

# Remove old objects so stale versions cannot be reused
rm(list = intersect(
  c(
    "supp2_source",
    "supp2_v2_dat",
    "supp2_v2_rows",
    "supp_table_2",
    "supp2_v2_raw_check"
  ),
  ls()
))

#-----------------------------
# Formatting helpers
#-----------------------------

fmt_n_pct <- function(n, d, digits = 1) {
  if (is.na(d) || d == 0) return("")
  sprintf(paste0("%d (%.", digits, "f)"), n, 100 * n / d)
}

fmt_pos_tested <- function(pos, tested, digits = 1) {
  if (is.na(tested) || tested == 0) return("")
  sprintf(paste0("%d/%d (%.", digits, "f)"), pos, tested, 100 * pos / tested)
}

fmt_total_n <- function(n) {
  sprintf("%d (100)", n)
}

fmt_total_tested <- function(n) {
  sprintf("/%d (100)", n)
}

#-----------------------------
# Use tab_dat if available
#-----------------------------
# Since your corrected Table 1 universe is now working, use tab_dat when it exists.
# Otherwise, fall back to dat and reapply the same filters.

supp2_source <- if (exists("tab_dat")) {
  tab_dat
} else {
  dat
}

if (!"sample_type" %in% names(supp2_source)) {
  supp2_source$sample_type <- supp2_source$substrate
}

#-----------------------------
# Build sample-level table universe
#-----------------------------

supp2_v2_dat <- supp2_source %>%
  mutate(
    unique_id = as.character(unique_id),
    site_id = as.character(site_id),
    carcass_id = as.character(carcass_id),
    carcass_association = as.character(carcass_association),
    sample_type = as.character(sample_type),
    substrate = as.character(substrate),
    
    time_since_death_years = suppressWarnings(as.numeric(time_since_death_years)),
    year_of_death = suppressWarnings(as.numeric(year_of_death)),
    distance_m = suppressWarnings(as.numeric(distance_m)),
    
    culture_done = as.character(culture_done),
    culture_ac_pos = as.character(culture_ac_pos),
    pcr_done = as.character(pcr_done),
    pcr_ac_pos = as.character(pcr_ac_pos),
    ac_any_pos = as.character(ac_any_pos),
    
    carcass_size_chr = as.character(carcass_size),
    carcass_size_num = suppressWarnings(as.numeric(carcass_size_chr)),
    habitat_class = as.character(habitat_class)
  ) %>%
  filter(
    carcass_association == "Y",
    sample_type %in% c("Tissue", "Soil"),
    !is.na(carcass_id),
    
    !str_detect(coalesce(unique_id, ""), "^WOLF-23-01"),
    !str_detect(coalesce(site_id, ""), "^WOLF-23-01"),
    !str_detect(coalesce(carcass_id, ""), "^WOLF-23-01")
  ) %>%
  mutate(
    culture_tested = culture_done == "Y" | culture_ac_pos %in% c("Y", "N"),
    pcr_tested = pcr_done == "Y" | pcr_ac_pos %in% c("Y", "N"),
    
    culture_pos = culture_ac_pos == "Y",
    pcr_pos = pcr_ac_pos == "Y",
    
    any_tested = culture_tested | pcr_tested,
    any_pos = case_when(
      ac_any_pos %in% c("Y", "N") ~ ac_any_pos == "Y",
      TRUE ~ culture_pos | pcr_pos
    ),
    
    tsd_cat = case_when(
      is.na(time_since_death_years) ~ "Unknown",
      time_since_death_years == 0 ~ "0 years",
      time_since_death_years == 1 ~ "1 year",
      time_since_death_years %% 1 == 0 ~ paste0(as.integer(time_since_death_years), " years"),
      TRUE ~ paste0(time_since_death_years, " years")
    ),
    
    death_year_cat = case_when(
      is.na(year_of_death) ~ "Unknown",
      TRUE ~ as.character(as.integer(year_of_death))
    ),
    
    carcass_size_cat = case_when(
      carcass_size_num == 2 ~ "Large",
      carcass_size_num == 1 ~ "Small",
      str_to_lower(carcass_size_chr) %in% c("large", "l") ~ "Large",
      str_to_lower(carcass_size_chr) %in% c("small", "s") ~ "Small",
      TRUE ~ "Unknown"
    ),
    
    habitat_moist_cat = case_when(
      habitat_class == "Tundra Barrens" ~ "Dry",
      habitat_class %in% c("Hummock Tundra", "Tundra Meadow") ~ "Medium",
      habitat_class %in% c("Riparian Tundra", "Riverbed", "Tundra Wetland") ~ "Wet",
      TRUE ~ "Unknown"
    ),
    
    distance_cat = case_when(
      sample_type != "Soil" ~ NA_character_,
      is.na(distance_m) ~ "Unknown",
      distance_m %% 1 == 0 ~ paste0(as.integer(distance_m), " m"),
      TRUE ~ paste0(distance_m, " m")
    )
  )

#-----------------------------
# Confirm table universe
#-----------------------------

supp2_v2_universe_check <- supp2_v2_dat %>%
  summarise(
    tissue_samples = sum(sample_type == "Tissue"),
    soil_samples = sum(sample_type == "Soil"),
    total_samples = n(),
    carcass_sites = n_distinct(carcass_id),
    culture_tested = sum(culture_tested, na.rm = TRUE),
    pcr_tested = sum(pcr_tested, na.rm = TRUE),
    any_tested = sum(any_tested, na.rm = TRUE)
  )

supp2_v2_universe_check

stopifnot(supp2_v2_universe_check$tissue_samples == 246)
stopifnot(supp2_v2_universe_check$soil_samples == 572)
stopifnot(supp2_v2_universe_check$total_samples == 818)
stopifnot(supp2_v2_universe_check$carcass_sites == 138)
stopifnot(supp2_v2_universe_check$culture_tested == 818)
stopifnot(supp2_v2_universe_check$pcr_tested == 669)
stopifnot(supp2_v2_universe_check$any_tested == 818)

#-----------------------------
# Raw diagnostic check
#-----------------------------
# This should show:
# 0 years: Tissue 84, Soil 27, Carcass sites 26
# 1 year:  Tissue 27, Soil 61, Carcass sites 29

supp2_v2_raw_check <- supp2_v2_dat %>%
  group_by(tsd_cat) %>%
  summarise(
    tissue_n = sum(sample_type == "Tissue"),
    soil_n = sum(sample_type == "Soil"),
    carcass_site_n = n_distinct(carcass_id),
    culture_pos_n = sum(culture_pos, na.rm = TRUE),
    culture_tested_n = sum(culture_tested, na.rm = TRUE),
    pcr_pos_n = sum(pcr_pos, na.rm = TRUE),
    pcr_tested_n = sum(pcr_tested, na.rm = TRUE),
    any_pos_n = sum(any_pos, na.rm = TRUE),
    any_tested_n = sum(any_tested, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(factor(
    tsd_cat,
    levels = c("0 years", "1 year", "2 years", "3 years", "5 years", "6 years", "Unknown")
  ))

supp2_v2_raw_check

#-----------------------------
# Denominators
#-----------------------------

total_tissue_samples <- sum(supp2_v2_dat$sample_type == "Tissue")
total_soil_samples <- sum(supp2_v2_dat$sample_type == "Soil")
total_carcass_sites <- n_distinct(supp2_v2_dat$carcass_id)

total_culture_tested <- sum(supp2_v2_dat$culture_tested, na.rm = TRUE)
total_pcr_tested <- sum(supp2_v2_dat$pcr_tested, na.rm = TRUE)
total_any_tested <- sum(supp2_v2_dat$any_tested, na.rm = TRUE)

#-----------------------------
# Row shell
#-----------------------------

supp2_v2_rows <- tribble(
  ~Variable, ~Category, ~category_column, ~soil_only,
  
  "Time Since Death", "0 years", "tsd_cat", FALSE,
  "Time Since Death", "1 year", "tsd_cat", FALSE,
  "Time Since Death", "2 years", "tsd_cat", FALSE,
  "Time Since Death", "3 years", "tsd_cat", FALSE,
  "Time Since Death", "5 years", "tsd_cat", FALSE,
  "Time Since Death", "6 years", "tsd_cat", FALSE,
  "Time Since Death", "Unknown", "tsd_cat", FALSE,
  
  "Year of Death", "2018", "death_year_cat", FALSE,
  "Year of Death", "2021", "death_year_cat", FALSE,
  "Year of Death", "2022", "death_year_cat", FALSE,
  "Year of Death", "2023", "death_year_cat", FALSE,
  "Year of Death", "2024", "death_year_cat", FALSE,
  "Year of Death", "Unknown", "death_year_cat", FALSE,
  
  "Carcass Size", "Large", "carcass_size_cat", FALSE,
  "Carcass Size", "Small", "carcass_size_cat", FALSE,
  "Carcass Size", "Unknown", "carcass_size_cat", FALSE,
  
  "Habitat Moisture Class", "Dry", "habitat_moist_cat", FALSE,
  "Habitat Moisture Class", "Medium", "habitat_moist_cat", FALSE,
  "Habitat Moisture Class", "Unknown", "habitat_moist_cat", FALSE,
  "Habitat Moisture Class", "Wet", "habitat_moist_cat", FALSE,
  
  "Soil Collection Distance", "0 m", "distance_cat", TRUE,
  "Soil Collection Distance", "1 m", "distance_cat", TRUE,
  "Soil Collection Distance", "3 m", "distance_cat", TRUE,
  "Soil Collection Distance", "5 m", "distance_cat", TRUE,
  "Soil Collection Distance", "15 m", "distance_cat", TRUE
)

#-----------------------------
# Calculate each displayed row independently
#-----------------------------

calc_supp2_row <- function(Variable, Category, category_column, soil_only) {
  
  row_dat <- supp2_v2_dat
  
  if (soil_only) {
    row_dat <- row_dat %>%
      filter(sample_type == "Soil")
  }
  
  row_dat <- row_dat %>%
    filter(.data[[category_column]] == Category)
  
  tissue_n <- sum(row_dat$sample_type == "Tissue", na.rm = TRUE)
  soil_n <- sum(row_dat$sample_type == "Soil", na.rm = TRUE)
  carcass_site_n <- n_distinct(row_dat$carcass_id)
  
  culture_pos_n <- sum(row_dat$culture_pos, na.rm = TRUE)
  culture_tested_n <- sum(row_dat$culture_tested, na.rm = TRUE)
  
  pcr_pos_n <- sum(row_dat$pcr_pos, na.rm = TRUE)
  pcr_tested_n <- sum(row_dat$pcr_tested, na.rm = TRUE)
  
  any_pos_n <- sum(row_dat$any_pos, na.rm = TRUE)
  any_tested_n <- sum(row_dat$any_tested, na.rm = TRUE)
  
  tibble(
    Variable = Variable,
    Category = Category,
    `Tissue Samples, n (%)` = if (soil_only) "-" else fmt_n_pct(tissue_n, total_tissue_samples),
    `Soil Samples, n (%)` = fmt_n_pct(soil_n, total_soil_samples),
    `Carcass Sites, n (%)` = fmt_n_pct(carcass_site_n, total_carcass_sites),
    `Culture + / n Tested (%)` = fmt_pos_tested(culture_pos_n, culture_tested_n),
    `PCR + / n Tested (%)` = fmt_pos_tested(pcr_pos_n, pcr_tested_n),
    `Any Test + / n Tested (%)` = fmt_pos_tested(any_pos_n, any_tested_n)
  )
}

#-----------------------------
# Final table
#-----------------------------

supp_table_2 <- pmap_dfr(
  supp2_v2_rows,
  calc_supp2_row
) %>%
  bind_rows(
    tibble(
      Variable = "",
      Category = "TOTAL",
      `Tissue Samples, n (%)` = fmt_total_n(total_tissue_samples),
      `Soil Samples, n (%)` = fmt_total_n(total_soil_samples),
      `Carcass Sites, n (%)` = fmt_total_n(total_carcass_sites),
      `Culture + / n Tested (%)` = fmt_pos_tested(
        sum(supp2_v2_dat$culture_pos, na.rm = TRUE),
        total_culture_tested
      ),
      `PCR + / n Tested (%)` = fmt_pos_tested(
        sum(supp2_v2_dat$pcr_pos, na.rm = TRUE),
        total_pcr_tested
      ),
      `Any Test + / n Tested (%)` = fmt_pos_tested(
        sum(supp2_v2_dat$any_pos, na.rm = TRUE),
        total_any_tested
      )
    )
  )

supp_table_2

#-----------------------------
# Export
#-----------------------------

knitr::kable(
  supp_table_2,
  caption = "Supplementary Table 2. Summary of tissue and soil samples stratified by model covariates."
)
```

## Delta Dataset Correlation Matrices
```{r, Soil Chemistry: Delta Dataset Correlation Matrices, cache=FALSE}

cor_pmat <- function(df, vars, method = "spearman") {
  pmat <- matrix(NA_real_, nrow = length(vars), ncol = length(vars),
                 dimnames = list(vars, vars))

  for (i in seq_along(vars)) {
    for (j in seq_along(vars)) {
      x <- df[[vars[i]]]
      y <- df[[vars[j]]]

      keep <- complete.cases(x, y)

      if (sum(keep) >= 3 &&
          stats::sd(x[keep]) > 0 &&
          stats::sd(y[keep]) > 0) {
        pmat[i, j] <- suppressWarnings(
          cor.test(x[keep], y[keep], method = method)$p.value
        )
      }
    }
  }

  pmat
}

# Keep variables with enough observed data + nonzero variance
delta_vars_abs_use <- delta_vars_abs[
  sapply(soil_delta_tsd[delta_vars_abs], function(x) sum(!is.na(x)) >= 3 && sd(x, na.rm = TRUE) > 0)
]

delta_vars_pct_use <- delta_vars_pct[
  sapply(soil_delta_tsd[delta_vars_pct], function(x) sum(!is.na(x)) >= 3 && sd(x, na.rm = TRUE) > 0)
]

# Absolute delta correlations
delta_abs_cor <- cor(
  soil_delta_tsd[, delta_vars_abs_use],
  use = "pairwise.complete.obs",
  method = "spearman"
)

delta_abs_p <- cor_pmat(
  soil_delta_tsd,
  vars = delta_vars_abs_use,
  method = "spearman"
)

round(delta_abs_cor, 2)
round(delta_abs_p, 3)

# Percentage delta correlations
delta_pct_cor <- cor(
  soil_delta_tsd[, delta_vars_pct_use],
  use = "pairwise.complete.obs",
  method = "spearman"
)

delta_pct_p <- cor_pmat(
  soil_delta_tsd,
  vars = delta_vars_pct_use,
  method = "spearman"
)

round(delta_pct_cor, 2)
round(delta_pct_p, 3)

plot_cor_heatmap <- function(cor_mat, p_mat = NULL, title = NULL) {

  cor_df <- as.data.frame(as.table(cor_mat)) %>%
    rename(
      Var1 = Var1,
      Var2 = Var2,
      correlation = Freq
    ) %>%
    mutate(
      Var1 = factor(Var1, levels = rownames(cor_mat)),
      Var2 = factor(Var2, levels = rev(colnames(cor_mat)))
    )

  if (!is.null(p_mat)) {
    p_df <- as.data.frame(as.table(p_mat)) %>%
      rename(
        Var1 = Var1,
        Var2 = Var2,
        p_value = Freq
      ) %>%
      mutate(
        sig_lab = case_when(
          is.na(p_value)      ~ "",
          p_value < 0.001     ~ "***",
          p_value < 0.01      ~ "**",
          p_value < 0.05      ~ "*",
          TRUE                ~ ""
        )
      )

    cor_df <- cor_df %>%
      left_join(p_df, by = c("Var1", "Var2"))
  } else {
    cor_df <- cor_df %>%
      mutate(sig_lab = "")
  }

  ggplot(cor_df, aes(x = Var1, y = Var2, fill = correlation)) +
    geom_tile(color = "white") +
    geom_text(aes(label = sprintf("%.2f%s", correlation, sig_lab)), size = 3) +
    scale_fill_gradient2(
      low = "#2166AC",
      mid = "white",
      high = "#B2182B",
      midpoint = 0,
      limits = c(-1, 1)
    ) +
    labs(
      title = title,
      x = NULL,
      y = NULL,
      fill = "r"
    ) +
    theme_minimal(base_size = 11) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      panel.grid = element_blank()
    )
}

# Absolute delta heatmap
p_delta_abs_cor <- plot_cor_heatmap(
  cor_mat = delta_abs_cor,
  p_mat   = delta_abs_p,
  title   = "Absolute Delta Soil Chemistry Correlations"
)

p_delta_abs_cor

# Percentage delta heatmap
p_delta_pct_cor <- plot_cor_heatmap(
  cor_mat = delta_pct_cor,
  p_mat   = delta_pct_p,
  title   = "Percentage Delta Soil Chemistry Correlations"
)

p_delta_pct_cor

```

## PERMANOVA Analyses WITHOUT Habitat Classes
```{r, Soil Chemistry: PERMANOVA Analyses WITHOUT Habitat Classes, cache=FALSE}

soil_perm4_tsd <- soil_chem_std %>%
  filter(time_since_death_years %in% c(2, 3)) %>%
  mutate(
    years_since_death_f = factor(
      paste0(time_since_death_years, "y"),
      levels = c("2y", "3y")
    ),
    grp_tsd_dist = factor(
      paste(years_since_death_f, distance_lab, sep = "_"),
      levels = c("2y_0m", "2y_15m", "3y_0m", "3y_15m")
    )
  )

soil_perm4_tsd %>% count(grp_tsd_dist)

res_perm4_tsd <- run_perm_suite(
  df         = soil_perm4_tsd,
  group_var  = "grp_tsd_dist",
  vars_z     = chem_vars_z,
  strata_var = "carcass_id",
  nperm      = 9999
)

res_perm4_tsd$permanova
res_perm4_tsd$permdisp
res_perm4_tsd$pairwise_permanova

```

## PERMANOVA Analyses WITH Habitat Classes
```{r, Soil Chemistry: PERMANOVA Analyses WITH Habitat Classes, cache=FALSE}

soil_perm12_tsd <- soil_chem_std %>%
  filter(time_since_death_years %in% c(2, 3), !is.na(habitat_moist3)) %>%
  mutate(
    years_since_death_f = factor(
      paste0(time_since_death_years, "y"),
      levels = c("2y", "3y")
    ),
    grp_tsd_dist_hab = factor(
      paste(years_since_death_f, distance_lab, habitat_moist3, sep = "_"),
      levels = c(
        "2y_0m_Dry", "2y_0m_Medium", "2y_0m_Wet",
        "2y_15m_Dry", "2y_15m_Medium", "2y_15m_Wet",
        "3y_0m_Dry", "3y_0m_Medium", "3y_0m_Wet",
        "3y_15m_Dry", "3y_15m_Medium", "3y_15m_Wet"
      )
    ) %>% droplevels()
  )

soil_perm12_tsd %>% count(grp_tsd_dist_hab)

res_perm12_tsd <- run_perm_suite(
  df         = soil_perm12_tsd,
  group_var  = "grp_tsd_dist_hab",
  vars_z     = chem_vars_z,
  strata_var = "carcass_id",
  nperm      = 9999
)

res_perm12_tsd$permanova
res_perm12_tsd$permdisp
res_perm12_tsd$pairwise_permanova

```

## PERMANOVA Absolute Delta Dataset
```{r, Soil Chemistry: PERMANOVA Absolute Delta Dataset, cache=FALSE}

soil_delta_abs_std <- soil_delta_tsd %>%
  drop_na(all_of(delta_vars_abs)) %>%
  mutate(
    across(all_of(delta_vars_abs), ~ as.numeric(scale(.x)), .names = "{.col}_z")
  )

delta_abs_z <- paste0(delta_vars_abs, "_z")

# 2-group: 2y vs 3y
res_perm_abs_tsd <- run_perm_suite(
  df         = soil_delta_abs_std,
  group_var  = "years_since_death_f",
  vars_z     = delta_abs_z,
  strata_var = "carcass_id",
  nperm      = 9999
)

res_perm_abs_tsd$permanova
res_perm_abs_tsd$permdisp

# 6-group: 2y/3y x habitat
soil_delta_abs_std <- soil_delta_abs_std %>%
  filter(!is.na(habitat_moist3)) %>%
  mutate(
    grp_tsd_hab = factor(
      paste(years_since_death_f, habitat_moist3, sep = "_"),
      levels = c(
        "2y_Dry", "2y_Medium", "2y_Wet",
        "3y_Dry", "3y_Medium", "3y_Wet"
      )
    )
  )

res_perm_abs_tsd_hab <- run_perm_suite(
  df         = soil_delta_abs_std,
  group_var  = "grp_tsd_hab",
  vars_z     = delta_abs_z,
  strata_var = "carcass_id",
  nperm      = 9999
)

res_perm_abs_tsd_hab$permanova
res_perm_abs_tsd_hab$permdisp
res_perm_abs_tsd_hab$pairwise_permanova

```

## PERMANOVA Percentage Delta Dataset
```{r, Soil Chemistry: PERMANOVA Percentage Delta Dataset, cache=FALSE}

soil_delta_pct_std <- soil_delta_tsd %>%
  drop_na(all_of(delta_vars_pct)) %>%
  mutate(
    across(all_of(delta_vars_pct), ~ as.numeric(scale(.x)), .names = "{.col}_z")
  )

delta_pct_z <- paste0(delta_vars_pct, "_z")

# 2-group: 2y vs 3y
res_perm_pct_tsd <- run_perm_suite(
  df         = soil_delta_pct_std,
  group_var  = "years_since_death_f",
  vars_z     = delta_pct_z,
  strata_var = "carcass_id",
  nperm      = 9999
)

res_perm_pct_tsd$permanova
res_perm_pct_tsd$permdisp

# 6-group: 2y/3y x habitat
soil_delta_pct_std <- soil_delta_pct_std %>%
  filter(!is.na(habitat_moist3)) %>%
  mutate(
    grp_tsd_hab = factor(
      paste(years_since_death_f, habitat_moist3, sep = "_"),
      levels = c(
        "2y_Dry", "2y_Medium", "2y_Wet",
        "3y_Dry", "3y_Medium", "3y_Wet"
      )
    )
  )

res_perm_pct_tsd_hab <- run_perm_suite(
  df         = soil_delta_pct_std,
  group_var  = "grp_tsd_hab",
  vars_z     = delta_pct_z,
  strata_var = "carcass_id",
  nperm      = 9999
)

res_perm_pct_tsd_hab$permanova
res_perm_pct_tsd_hab$permdisp
res_perm_pct_tsd_hab$pairwise_permanova

```

## Paired Tests Delta Datasets
```{r, Soil Chemistry: Paired Tests Delta Datasets, cache=FALSE}

library(DT)

mean_or_na <- function(x) {
  if (all(is.na(x))) NA_real_ else mean(x, na.rm = TRUE)
}

sd_or_na <- function(x) {
  x <- x[!is.na(x)]
  if (length(x) < 2) NA_real_ else sd(x)
}

mean_diff_ci <- function(diffs, conf.level = 0.95) {
  diffs <- diffs[!is.na(diffs)]
  n <- length(diffs)

  if (n < 2) {
    return(tibble(
      mean_diff = ifelse(n == 1, mean(diffs), NA_real_),
      diff_ci_low = NA_real_,
      diff_ci_high = NA_real_
    ))
  }

  md <- mean(diffs)
  se <- sd(diffs) / sqrt(n)
  tcrit <- qt(1 - (1 - conf.level) / 2, df = n - 1)

  tibble(
    mean_diff = md,
    diff_ci_low = md - tcrit * se,
    diff_ci_high = md + tcrit * se
  )
}

run_paired_test_suite <- function(df, outcome_vars) {

  purrr::map_dfr(outcome_vars, function(v) {

    dwide <- df %>%
      dplyr::select(carcass_id, years_since_death_f, dplyr::all_of(v)) %>%
      dplyr::filter(years_since_death_f %in% c("2y", "3y")) %>%
      dplyr::filter(!is.na(.data[[v]])) %>%
      dplyr::distinct() %>%
      tidyr::pivot_wider(
        names_from = years_since_death_f,
        values_from = dplyr::all_of(v)
      ) %>%
      tidyr::drop_na(`2y`, `3y`)

    n_pairs <- nrow(dwide)

    diffs <- if (n_pairs > 0) dwide$`3y` - dwide$`2y` else numeric(0)

    ci_tbl <- mean_diff_ci(diffs)

    out_base <- tibble(
      variable = v,
      n_pairs = n_pairs,

      mean_2y = if (n_pairs > 0) mean(dwide$`2y`, na.rm = TRUE) else NA_real_,
      sd_2y   = if (n_pairs > 0) sd_or_na(dwide$`2y`) else NA_real_,

      mean_3y = if (n_pairs > 0) mean(dwide$`3y`, na.rm = TRUE) else NA_real_,
      sd_3y   = if (n_pairs > 0) sd_or_na(dwide$`3y`) else NA_real_,

      mean_diff_3y_minus_2y = ci_tbl$mean_diff,
      diff_ci_low = ci_tbl$diff_ci_low,
      diff_ci_high = ci_tbl$diff_ci_high
    )

    if (n_pairs < 3) {
      return(
        out_base %>%
          dplyr::mutate(
            normality_p = NA_real_,
            test_used   = NA_character_,
            statistic   = NA_real_,
            p           = NA_real_
          )
      )
    }

    shapiro_p <- if (sd(diffs, na.rm = TRUE) > 0) {
      shapiro.test(diffs)$p.value
    } else {
      NA_real_
    }

    if (!is.na(shapiro_p) && shapiro_p > 0.05) {
      tt <- t.test(dwide$`3y`, dwide$`2y`, paired = TRUE)

      out_base %>%
        dplyr::mutate(
          normality_p = shapiro_p,
          test_used   = "Paired t-test",
          statistic   = unname(tt$statistic),
          p           = tt$p.value
        )

    } else {
      wt <- wilcox.test(dwide$`3y`, dwide$`2y`, paired = TRUE, exact = FALSE)

      out_base %>%
        dplyr::mutate(
          normality_p = shapiro_p,
          test_used   = "Wilcoxon signed-rank",
          statistic   = unname(wt$statistic),
          p           = wt$p.value
        )
    }
  }) %>%
    dplyr::mutate(p_adj_BH = p.adjust(p, method = "BH")) %>%
    dplyr::arrange(p_adj_BH, p)
}

paired_tests_abs <- run_paired_test_suite(
  df = soil_delta_tsd,
  outcome_vars = delta_vars_abs
)

paired_tests_pct <- run_paired_test_suite(
  df = soil_delta_tsd,
  outcome_vars = delta_vars_pct
)

DT::datatable(
  paired_tests_abs,
  options = list(pageLength = 25, scrollX = TRUE),
  caption = "Paired 2y vs 3y tests: absolute delta dataset"
)

DT::datatable(
  paired_tests_pct,
  options = list(pageLength = 25, scrollX = TRUE),
  caption = "Paired 2y vs 3y tests: percentage delta dataset"
)

```

## Paired Tests, 0m vs 15m within 2y
```{r, Soil Chemistry: Paired Tests, 0m vs 15m within 2y, cache=FALSE}

mean_or_na <- function(x) {
  if (all(is.na(x))) NA_real_ else mean(x, na.rm = TRUE)
}

run_paired_distance_tests_2y <- function(df, vars) {

  wide_df <- df %>%
    dplyr::filter(
      time_since_death_years == 2,
      distance_lab %in% c("0m", "15m")
    ) %>%
    dplyr::group_by(carcass_id, distance_lab) %>%
    dplyr::summarise(
      dplyr::across(dplyr::all_of(vars), mean_or_na),
      .groups = "drop"
    ) %>%
    tidyr::pivot_wider(
      names_from = distance_lab,
      values_from = dplyr::all_of(vars),
      names_glue = "{.value}_{distance_lab}"
    )

  purrr::map_dfr(vars, function(v) {

    v0  <- paste0(v, "_0m")
    v15 <- paste0(v, "_15m")

    dsub <- wide_df %>%
      dplyr::select(carcass_id, dplyr::all_of(c(v0, v15))) %>%
      tidyr::drop_na()

    n_pairs <- nrow(dsub)

    diffs <- if (n_pairs > 0) dsub[[v0]] - dsub[[v15]] else numeric(0)

    ci_tbl <- mean_diff_ci(diffs)

    out_base <- tibble(
      variable = v,
      n_pairs = n_pairs,

      mean_0m = if (n_pairs > 0) mean(dsub[[v0]], na.rm = TRUE) else NA_real_,
      sd_0m   = if (n_pairs > 0) sd_or_na(dsub[[v0]]) else NA_real_,

      mean_15m = if (n_pairs > 0) mean(dsub[[v15]], na.rm = TRUE) else NA_real_,
      sd_15m   = if (n_pairs > 0) sd_or_na(dsub[[v15]]) else NA_real_,

      mean_diff_0m_minus_15m = ci_tbl$mean_diff,
      diff_ci_low = ci_tbl$diff_ci_low,
      diff_ci_high = ci_tbl$diff_ci_high
    )

    if (n_pairs < 3) {
      return(
        out_base %>%
          dplyr::mutate(
            normality_p = NA_real_,
            test_used = NA_character_,
            statistic = NA_real_,
            p = NA_real_
          )
      )
    }

    shapiro_p <- if (sd(diffs, na.rm = TRUE) > 0) {
      shapiro.test(diffs)$p.value
    } else {
      NA_real_
    }

    if (!is.na(shapiro_p) && shapiro_p > 0.05) {
      tt <- t.test(dsub[[v0]], dsub[[v15]], paired = TRUE)

      out_base %>%
        dplyr::mutate(
          normality_p = shapiro_p,
          test_used = "Paired t-test",
          statistic = unname(tt$statistic),
          p = tt$p.value
        )

    } else {
      wt <- wilcox.test(dsub[[v0]], dsub[[v15]], paired = TRUE, exact = FALSE)

      out_base %>%
        dplyr::mutate(
          normality_p = shapiro_p,
          test_used = "Wilcoxon signed-rank",
          statistic = unname(wt$statistic),
          p = wt$p.value
        )
    }
  }) %>%
    dplyr::mutate(p_adj_BH = p.adjust(p, method = "BH")) %>%
    dplyr::arrange(p_adj_BH, p)
}

paired_tests_2y_0m_vs_15m <- run_paired_distance_tests_2y(
  df = soil_chem_tr,
  vars = chem_vars_use
)

DT::datatable(
  paired_tests_2y_0m_vs_15m,
  options = list(pageLength = 25, scrollX = TRUE),
  caption = "Paired 0m vs 15m tests within the 2-year-after-death cohort"
)

paired_tests_2y_0m_vs_15m %>%
  mutate(across(where(is.numeric), ~ round(.x, 3))) %>%
  knitr::kable(
    caption = "Paired 0m vs 15m tests within the 2-year-after-death cohort"
  )
```

## GLMMs on Individual Delta Soil Properties
```{r, Soil Chemistry: GLMMs on Individual Delta Soil Properties, cache=FALSE}

library(lmerTest)
library(broom.mixed)

fit_delta_mixed_models <- function(df, outcome_vars) {

  model_list <- map(outcome_vars, function(v) {

    dsub <- df %>%
      filter(
        years_since_death_f %in% c("2y", "3y"),
        !is.na(.data[[v]]),
        !is.na(habitat_moist3),
        !is.na(carcass_id)
      ) %>%
      droplevels()

    if (nrow(dsub) < 6 || dplyr::n_distinct(dsub$carcass_id) < 3) {
      return(list(
        fit = NULL,
        tidy = tibble(
          variable = v,
          term = NA_character_,
          estimate = NA_real_,
          std.error = NA_real_,
          statistic = NA_real_,
          p.value = NA_real_,
          conf.low = NA_real_,
          conf.high = NA_real_,
          n_obs = nrow(dsub),
          n_carcass = dplyr::n_distinct(dsub$carcass_id),
          AIC = NA_real_
        )
      ))
    }

    form <- as.formula(
      paste0(v, " ~ years_since_death_f + habitat_moist3 + (1 | carcass_id)")
    )

    fit <- glmmTMB(
      formula = form,
      data = dsub,
      family = gaussian()
    )

    td <- broom.mixed::tidy(
      fit,
      effects = "fixed",
      conf.int = TRUE
    ) %>%
      mutate(
        variable = v,
        n_obs = nrow(dsub),
        n_carcass = dplyr::n_distinct(dsub$carcass_id),
        AIC = AIC(fit)
      ) %>%
      select(variable, term, estimate, std.error, statistic, p.value,
             conf.low, conf.high, n_obs, n_carcass, AIC)

    list(fit = fit, tidy = td)
  })

  names(model_list) <- outcome_vars

  tab <- bind_rows(map(model_list, "tidy"))

  list(
    models = model_list,
    table = tab
  )
}

abs_delta_model_fits <- fit_delta_mixed_models(
  df = soil_delta_tsd,
  outcome_vars = delta_vars_abs
)

pct_delta_model_fits <- fit_delta_mixed_models(
  df = soil_delta_tsd,
  outcome_vars = delta_vars_pct
)

abs_delta_model_fits$table
pct_delta_model_fits$table

knitr::kable(abs_delta_model_fits$table, digits = 3,
             caption = "Time and Habitat mixed models: absolute delta dataset")
knitr::kable(pct_delta_model_fits$table, digits = 3,
             caption = "Time and Habitat mixed models: percentage delta dataset")

### Habitat-Only Models ###

fit_delta_habitat_models <- function(df, outcome_vars) {

  model_list <- purrr::map(outcome_vars, function(v) {

    dsub <- df %>%
      filter(
        !is.na(.data[[v]]),
        !is.na(habitat_moist3),
        !is.na(carcass_id)
      ) %>%
      droplevels()

    if (nrow(dsub) < 6 || dplyr::n_distinct(dsub$carcass_id) < 3) {
      return(list(
        fit = NULL,
        tidy = tibble::tibble(
          variable = v,
          term = NA_character_,
          estimate = NA_real_,
          std.error = NA_real_,
          statistic = NA_real_,
          p.value = NA_real_,
          conf.low = NA_real_,
          conf.high = NA_real_,
          n_obs = nrow(dsub),
          n_carcass = dplyr::n_distinct(dsub$carcass_id),
          singular = NA
        )
      ))
    }

    form <- as.formula(
      paste0(v, " ~ habitat_moist3 + (1 | carcass_id)")
    )

    fit <- lmerTest::lmer(
      form,
      data = dsub,
      REML = FALSE,
      control = lme4::lmerControl(
        optimizer = "bobyqa",
        optCtrl = list(maxfun = 1e5)
      )
    )

    td <- broom.mixed::tidy(
      fit,
      effects = "fixed",
      conf.int = TRUE
    ) %>%
      mutate(
        variable = v,
        n_obs = nrow(dsub),
        n_carcass = dplyr::n_distinct(dsub$carcass_id),
        singular = lme4::isSingular(fit, tol = 1e-4)
      ) %>%
      select(variable, term, estimate, std.error, statistic, p.value,
             conf.low, conf.high, n_obs, n_carcass, singular)

    list(fit = fit, tidy = td)
  })

  names(model_list) <- outcome_vars

  list(
    models = model_list,
    table = bind_rows(purrr::map(model_list, "tidy"))
  )
}

abs_hab_models <- fit_delta_habitat_models(
  df = soil_delta_tsd,
  outcome_vars = delta_vars_abs
)

pct_hab_models <- fit_delta_habitat_models(
  df = soil_delta_tsd,
  outcome_vars = delta_vars_pct
)

knitr::kable(abs_hab_models$table, digits = 3,
             caption = "Habitat-only mixed models: absolute delta dataset")
knitr::kable(pct_hab_models$table, digits = 3,
             caption = "Habitat-only mixed models: percentage delta dataset")

```


## Soil Moisture Visualizations
```{r, Soil Chemistry: Soil Moisture Visualizations, cache=FALSE}
# Raw paired soil moisture: 0m and 15m shown separately
soil_moist_raw_plot <- soil_chem_tr %>%
  filter(
    time_since_death_years %in% c(2, 3),
    distance_lab %in% c("0m", "15m"),
    !is.na(soil_moisture),
    !is.na(habitat_moist3)
  ) %>%
  mutate(
    years_since_death_f = factor(
      paste0(time_since_death_years, "y"),
      levels = c("2y", "3y")
    )
  )

p_soil_moist_raw <- ggplot(
  soil_moist_raw_plot,
  aes(
    x = habitat_moist3,
    y = soil_moisture,
    color = distance_lab,
    shape = years_since_death_f
  )
) +
  geom_point(
    position = position_jitterdodge(jitter.width = 0.15, dodge.width = 0.65),
    size = 2.7,
    alpha = 0.85
  ) +
  stat_summary(
    aes(group = distance_lab),
    fun = median,
    geom = "crossbar",
    width = 0.45,
    position = position_dodge(width = 0.65),
    alpha = 0.6
  ) +
  labs(
    title = "Raw soil moisture by habitat moisture class",
    subtitle = "0m carcass-center and 15m paired-control soils shown separately",
    x = "Habitat moisture class",
    y = "Soil moisture",
    color = "Distance",
    shape = "Years since death"
  ) +
  theme_bw() +
  theme(
    panel.grid.minor = element_blank()
  )

p_soil_moist_raw


# Absolute and percentage delta soil moisture
soil_moist_delta_plot <- soil_delta_tsd %>%
  filter(!is.na(habitat_moist3)) %>%
  select(
    carcass_id,
    years_since_death_f,
    habitat_moist3,
    delta_soil_moisture,
    pct_delta_soil_moisture
  ) %>%
  pivot_longer(
    cols = c(delta_soil_moisture, pct_delta_soil_moisture),
    names_to = "delta_metric",
    values_to = "soil_moisture_delta"
  ) %>%
  mutate(
    delta_metric = recode(
      delta_metric,
      delta_soil_moisture = "Absolute delta: 0m - 15m",
      pct_delta_soil_moisture = "Percentage delta: ((0m - 15m) / 15m) * 100"
    )
  )

p_soil_moist_delta <- ggplot(
  soil_moist_delta_plot,
  aes(
    x = habitat_moist3,
    y = soil_moisture_delta,
    color = years_since_death_f,
    shape = years_since_death_f
  )
) +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.6) +
  geom_point(
    position = position_jitterdodge(jitter.width = 0.15, dodge.width = 0.55),
    size = 2.7,
    alpha = 0.85
  ) +
  stat_summary(
    aes(group = years_since_death_f),
    fun = median,
    geom = "crossbar",
    width = 0.45,
    position = position_dodge(width = 0.55),
    alpha = 0.6
  ) +
  facet_wrap(~ delta_metric, scales = "free_y") +
  labs(
    title = "Carcass-associated soil moisture deltas by habitat moisture class",
    subtitle = "Positive values indicate greater soil moisture at 0m than at 15m",
    x = "Habitat moisture class",
    y = "Soil moisture delta",
    color = "Years since death",
    shape = "Years since death"
  ) +
  theme_bw() +
  theme(
    panel.grid.minor = element_blank()
  )

p_soil_moist_delta

```


## Soil Water Holding Capacity Visualizations
```{r, Soil Chemistry: Soil Water Holding Capacity Visualizations, cache=FALSE}

# Raw paired water holding capacity: 0m and 15m shown separately
water_hold_raw_plot <- soil_chem_tr %>%
  filter(
    time_since_death_years %in% c(2, 3),
    distance_lab %in% c("0m", "15m"),
    !is.na(soil_water_hold_cap_log),
    !is.na(habitat_moist3)
  ) %>%
  mutate(
    years_since_death_f = factor(
      paste0(time_since_death_years, "y"),
      levels = c("2y", "3y")
    )
  )

p_water_hold_raw <- ggplot(
  water_hold_raw_plot,
  aes(
    x = habitat_moist3,
    y = soil_water_hold_cap_log,
    color = distance_lab,
    shape = years_since_death_f
  )
) +
  geom_point(
    position = position_jitterdodge(jitter.width = 0.15, dodge.width = 0.65),
    size = 2.7,
    alpha = 0.85
  ) +
  stat_summary(
    aes(group = distance_lab),
    fun = median,
    geom = "crossbar",
    width = 0.45,
    position = position_dodge(width = 0.65),
    alpha = 0.6
  ) +
  labs(
    title = "Raw water holding capacity by habitat moisture class",
    subtitle = "0m carcass-center and 15m paired-control soils shown separately",
    x = "Habitat moisture class",
    y = "Water holding capacity, log-transformed",
    color = "Distance",
    shape = "Years since death"
  ) +
  theme_bw() +
  theme(
    panel.grid.minor = element_blank()
  )

p_water_hold_raw


# Absolute and percentage delta water holding capacity
water_hold_delta_plot <- soil_delta_tsd %>%
  filter(!is.na(habitat_moist3)) %>%
  select(
    carcass_id,
    years_since_death_f,
    habitat_moist3,
    delta_soil_water_hold_cap_log,
    pct_delta_soil_water_hold_cap_log
  ) %>%
  pivot_longer(
    cols = c(delta_soil_water_hold_cap_log, pct_delta_soil_water_hold_cap_log),
    names_to = "delta_metric",
    values_to = "water_hold_delta"
  ) %>%
  mutate(
    delta_metric = recode(
      delta_metric,
      delta_soil_water_hold_cap_log = "Absolute delta: 0m - 15m",
      pct_delta_soil_water_hold_cap_log = "Percentage delta: ((0m - 15m) / 15m) * 100"
    )
  )

p_water_hold_delta <- ggplot(
  water_hold_delta_plot,
  aes(
    x = habitat_moist3,
    y = water_hold_delta,
    color = years_since_death_f,
    shape = years_since_death_f
  )
) +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.6) +
  geom_point(
    position = position_jitterdodge(jitter.width = 0.15, dodge.width = 0.55),
    size = 2.7,
    alpha = 0.85
  ) +
  stat_summary(
    aes(group = years_since_death_f),
    fun = median,
    geom = "crossbar",
    width = 0.45,
    position = position_dodge(width = 0.55),
    alpha = 0.6
  ) +
  facet_wrap(~ delta_metric, scales = "free_y") +
  labs(
    title = "Carcass-associated water holding capacity deltas by habitat moisture class",
    subtitle = "Positive values indicate greater water holding capacity at 0m than at 15m",
    x = "Habitat moisture class",
    y = "Water holding capacity delta",
    color = "Years since death",
    shape = "Years since death"
  ) +
  theme_bw() +
  theme(
    panel.grid.minor = element_blank()
  )

p_water_hold_delta

```


## LASSO and RIDGE Viability Models
### Data Setup
```{r, Soil Chemistry: Penalized Viability Models Setup, cache=FALSE}
library(glmmLasso)
library(hglm)

first_nonmissing <- function(x) {
  x <- x[!is.na(x)]
  if (length(x) == 0) NA else x[1]
}

mean_or_na <- function(x) {
  if (all(is.na(x))) NA_real_ else mean(x, na.rm = TRUE)
}

scale_safe <- function(x) {
  if (sum(!is.na(x)) < 2 || sd(x, na.rm = TRUE) == 0) {
    rep(NA_real_, length(x))
  } else {
    as.numeric(scale(x))
  }
}

# Rebuild raw soil chemistry viability dataset at the carcass-time-distance level
# keeps 0m and 15m as separate observations
soil_raw_viab <- soil_chem_tr %>%
  filter(
    time_since_death_years %in% c(2, 3),
    distance_lab %in% c("0m", "15m")
  ) %>%
  mutate(
    y_cult_row = case_when(
      culture_done == "Y" & culture_ac_pos == "Y" ~ 1L,
      culture_done == "Y" & culture_ac_pos != "Y" ~ 0L,
      TRUE ~ NA_integer_
    ),
    years_since_death_f = factor(
      paste0(time_since_death_years, "y"),
      levels = c("2y", "3y")
    ),
    carcass_size = case_when(
      as.character(carcass_size) %in% c("1", "Small") ~ "Small",
      as.character(carcass_size) %in% c("2", "Large") ~ "Large",
      TRUE ~ NA_character_
    ),
    carcass_size = factor(carcass_size, levels = c("Small", "Large")),
    habitat_moist3 = factor(habitat_moist3, levels = c("Dry", "Medium", "Wet")),
    distance_lab = factor(distance_lab, levels = c("0m", "15m"))
  ) %>%
  group_by(carcass_id, time_since_death_years, years_since_death_f, distance_lab) %>%
  summarise(
    y_cult = if_else(
      any(!is.na(y_cult_row)),
      as.integer(any(y_cult_row == 1, na.rm = TRUE)),
      NA_integer_
    ),
    across(all_of(chem_vars_use), mean_or_na),
    habitat_moist3 = first_nonmissing(habitat_moist3),
    carcass_size = first_nonmissing(carcass_size),
    .groups = "drop"
  ) %>%
  mutate(
    carcass_id = factor(carcass_id),
    habitat_moist3 = factor(habitat_moist3, levels = c("Dry", "Medium", "Wet")),
    carcass_size = factor(carcass_size, levels = c("Small", "Large")),
    years_since_death_f = factor(years_since_death_f, levels = c("2y", "3y")),
    distance_lab = factor(distance_lab, levels = c("0m", "15m"))
  ) %>%
  mutate(
    across(all_of(chem_vars_use), scale_safe, .names = "{.col}_pen_z")
  )

raw_chem_pen_vars <- paste0(chem_vars_use, "_pen_z")

# For delta datasets, use 0m viability as the outcome
viab_0m <- soil_raw_viab %>%
  filter(distance_lab == "0m") %>%
  dplyr::select(
    carcass_id,
    time_since_death_years,
    y_cult,
    carcass_size
  )

soil_delta_viab <- soil_delta_tsd %>%
  left_join(
    viab_0m,
    by = c("carcass_id", "time_since_death_years")
  ) %>%
  mutate(
    carcass_id = factor(carcass_id),
    habitat_moist3 = factor(habitat_moist3, levels = c("Dry", "Medium", "Wet")),
    carcass_size = factor(carcass_size, levels = c("Small", "Large")),
    years_since_death_f = factor(years_since_death_f, levels = c("2y", "3y"))
  )

# Standardized absolute delta predictors
soil_delta_abs_viab <- soil_delta_viab %>%
  mutate(
    across(all_of(delta_vars_abs), scale_safe, .names = "{.col}_pen_z")
  )

delta_abs_pen_vars <- paste0(delta_vars_abs, "_pen_z")

# Standardized percentage delta predictors
soil_delta_pct_viab <- soil_delta_viab %>%
  mutate(
    across(all_of(delta_vars_pct), scale_safe, .names = "{.col}_pen_z")
  )

delta_pct_pen_vars <- paste0(delta_vars_pct, "_pen_z")

#Balance checks
check_viability_balance <- function(df, label) {
  df %>%
    summarise(
      dataset = label,
      n = n(),
      n_carcass = n_distinct(carcass_id),
      positives = sum(y_cult == 1, na.rm = TRUE),
      negatives = sum(y_cult == 0, na.rm = TRUE),
      missing_y = sum(is.na(y_cult))
    )
}

bind_rows(
  check_viability_balance(soil_raw_viab, "Raw standardized chemistry"),
  check_viability_balance(soil_delta_abs_viab, "Absolute delta chemistry"),
  check_viability_balance(soil_delta_pct_viab, "Percentage delta chemistry")
) %>%
  knitr::kable(caption = "Viability outcome balance for penalized regression datasets")
```

### Helper Functions
```{r, Soil Chemistry: Penalized Viability Models Helper Functions, cache=FALSE}
prep_pen_model_data <- function(df, outcome, predictors, group_var = "carcass_id") {

  df2 <- df %>%
    dplyr::select(all_of(c(outcome, group_var, predictors))) %>%
    drop_na()

  df2[[outcome]] <- as.integer(df2[[outcome]])
  df2[[group_var]] <- factor(df2[[group_var]])

  if (dplyr::n_distinct(df2[[outcome]]) < 2) {
    stop("Outcome has fewer than two classes after filtering.")
  }

  df2
}


fit_glmmLasso_grid <- function(df,
                               predictors,
                               outcome = "y_cult",
                               group_var = "carcass_id",
                               lambda_grid = c(10, 25, 50, 100, 200, 500, 1000)) {

  df2 <- prep_pen_model_data(
    df = df,
    outcome = outcome,
    predictors = predictors,
    group_var = group_var
  )

  fix_form <- stats::reformulate(predictors, response = outcome)

  rnd_form <- list(~ 1)
  names(rnd_form) <- group_var

  grid_fits <- purrr::map_dfr(lambda_grid, function(lam) {

    warning_messages <- character(0)

    fit <- withCallingHandlers(
      tryCatch(
        glmmLasso::glmmLasso(
          fix = fix_form,
          rnd = rnd_form,
          data = df2,
          lambda = lam,
          family = stats::binomial(link = "logit"),
          switch.NR = TRUE,
          final.re = FALSE,
          control = list(
            print.iter = FALSE,
            print.iter.final = FALSE,
            maxIter = 100,
            maxIter.final = 100,
            epsilon = 1e-4
          )
        ),
        error = function(e) e
      ),
      warning = function(w) {
        warning_messages <<- c(warning_messages, conditionMessage(w))
        invokeRestart("muffleWarning")
      }
    )

    if (inherits(fit, "error")) {
      tibble::tibble(
        lambda = lam,
        aic = NA_real_,
        bic = NA_real_,
        conv_step = NA_real_,
        n_selected = NA_integer_,
        convergence_warning = TRUE,
        warnings = conditionMessage(fit),
        fit = list(NULL)
      )
    } else {

      coefs <- as.numeric(fit$coefficients)
      names(coefs) <- names(fit$coefficients)

      tibble::tibble(
        lambda = lam,
        aic = as.numeric(fit$aic),
        bic = as.numeric(fit$bic),
        conv_step = as.numeric(fit$conv.step),
        n_selected = sum(abs(coefs[names(coefs) != "(Intercept)"]) > 1e-6, na.rm = TRUE),
        convergence_warning = any(grepl("did not converge|Algorithm did not converge|Fisher", warning_messages)),
        warnings = paste(unique(warning_messages), collapse = " | "),
        fit = list(fit)
      )
    }
  })

  valid_fits <- grid_fits %>%
    dplyr::filter(!is.na(.data$bic))

  if (nrow(valid_fits) == 0) {
    stop("No glmmLasso models fit successfully across the lambda grid.")
  }

  # Prefer models without convergence warnings; if all warned, choose lowest BIC anyway
  best_row <- valid_fits %>%
    dplyr::arrange(.data$convergence_warning, .data$bic) %>%
    dplyr::slice(1)

  best_fit <- best_row$fit[[1]]

  coef_table <- tibble::tibble(
    term = names(best_fit$coefficients),
    estimate = as.numeric(best_fit$coefficients)
  ) %>%
    dplyr::mutate(
      selected = abs(.data$estimate) > 1e-6
    ) %>%
    dplyr::arrange(dplyr::desc(.data$selected), dplyr::desc(abs(.data$estimate)))

  list(
    data = df2,
    grid = grid_fits %>% dplyr::select(-.data$fit),
    best_lambda = best_row$lambda,
    best_fit = best_fit,
    coef_table = coef_table
  )
}


fit_hglm_ridge <- function(df,
                           predictors,
                           outcome = "y_cult",
                           group_var = "carcass_id") {

  df2 <- prep_pen_model_data(
    df = df,
    outcome = outcome,
    predictors = predictors,
    group_var = group_var
  )

  # Fixed intercept only
  X_fixed <- matrix(1, nrow = nrow(df2), ncol = 1)
  colnames(X_fixed) <- "(Intercept)"

  Z_carcass <- model.matrix(
    reformulate(group_var, intercept = FALSE),
    data = df2
  )

  Z_pred <- model.matrix(
    reformulate(predictors),
    data = df2
  )

  Z_pred <- Z_pred[, colnames(Z_pred) != "(Intercept)", drop = FALSE]

  # Remove zero-variance predictor columns after dummy expansion
  keep_cols <- apply(Z_pred, 2, function(z) sd(z, na.rm = TRUE) > 0)
  Z_pred <- Z_pred[, keep_cols, drop = FALSE]

  Z_all <- cbind(Z_carcass, Z_pred)

  fit <- hglm::hglm(
    X = X_fixed,
    y = df2[[outcome]],
    Z = Z_all,
    family = binomial(link = "logit"),
    rand.family = list(
      gaussian(link = "identity"),
      gaussian(link = "identity")
    ),
    RandC = c(ncol(Z_carcass), ncol(Z_pred)),
    fix.disp = 1,
    method = "EQL1",
    maxit = 100,
    sparse = TRUE,
    calc.like = TRUE
  )

  ranef_vec <- as.numeric(fit$ranef)

  ridge_coef <- tibble(
    term = colnames(Z_pred),
    ridge_estimate = tail(ranef_vec, ncol(Z_pred))
  ) %>%
    arrange(desc(abs(ridge_estimate)))

  list(
    data = df2,
    fit = fit,
    coef_table = ridge_coef,
    convergence = fit$Converge,
    var_components = fit$varRanef
  )
}


summarise_penalized_results <- function(lasso_obj, ridge_obj, label) {

  lasso_tbl <- lasso_obj$coef_table %>%
    filter(term != "(Intercept)") %>%
    transmute(
      dataset = label,
      term,
      lasso_estimate = estimate,
      lasso_selected = selected
    )

  ridge_tbl <- ridge_obj$coef_table %>%
    transmute(
      term,
      ridge_estimate
    )

  full_join(lasso_tbl, ridge_tbl, by = "term") %>%
    arrange(desc(lasso_selected), desc(abs(ridge_estimate)))
}

```

### Models
```{r, Soil Chemistry: Penalized Viability Models, cache=FALSE}

# Raw chemistry model parameters
raw_pen_predictors <- c(
  raw_chem_pen_vars,
  "habitat_moist3",
  "years_since_death_f",
  "carcass_size",
  "distance_lab"
)

# Absolute delta model parameters
abs_pen_predictors <- c(
  delta_abs_pen_vars,
  "habitat_moist3",
  "years_since_death_f",
  "carcass_size"
)

# Percentage delta model parameters
pct_pen_predictors <- c(
  delta_pct_pen_vars,
  "habitat_moist3",
  "years_since_death_f",
  "carcass_size"
)


### Raw Chemistry Models ###

raw_lasso <- fit_glmmLasso_grid(
  df = soil_raw_viab,
  predictors = raw_pen_predictors,
  outcome = "y_cult",
  group_var = "carcass_id"
)

raw_ridge <- fit_hglm_ridge(
  df = soil_raw_viab,
  predictors = raw_pen_predictors,
  outcome = "y_cult",
  group_var = "carcass_id"
)

raw_lasso$grid %>%
  knitr::kable(digits = 3, caption = "glmmLASSO tuning grid: raw standardized chemistry")

raw_pen_results <- summarise_penalized_results(
  lasso_obj = raw_lasso,
  ridge_obj = raw_ridge,
  label = "Raw standardized chemistry"
)

DT::datatable(
  raw_pen_results,
  options = list(pageLength = 30, scrollX = TRUE),
  caption = "Penalized viability model results: raw standardized chemistry"
)


### Absolute Delta Models ###

abs_lasso <- fit_glmmLasso_grid(
  df = soil_delta_abs_viab,
  predictors = abs_pen_predictors,
  outcome = "y_cult",
  group_var = "carcass_id"
)

abs_ridge <- fit_hglm_ridge(
  df = soil_delta_abs_viab,
  predictors = abs_pen_predictors,
  outcome = "y_cult",
  group_var = "carcass_id"
)

abs_lasso$grid %>%
  knitr::kable(digits = 3, caption = "glmmLASSO tuning grid: absolute delta chemistry")

abs_pen_results <- summarise_penalized_results(
  lasso_obj = abs_lasso,
  ridge_obj = abs_ridge,
  label = "Absolute delta chemistry"
)

DT::datatable(
  abs_pen_results,
  options = list(pageLength = 30, scrollX = TRUE),
  caption = "Penalized viability model results: absolute delta chemistry"
)


### Percentage Delta Models ###

pct_lasso <- fit_glmmLasso_grid(
  df = soil_delta_pct_viab,
  predictors = pct_pen_predictors,
  outcome = "y_cult",
  group_var = "carcass_id"
)

pct_ridge <- fit_hglm_ridge(
  df = soil_delta_pct_viab,
  predictors = pct_pen_predictors,
  outcome = "y_cult",
  group_var = "carcass_id"
)

pct_lasso$grid %>%
  knitr::kable(digits = 3, caption = "glmmLASSO tuning grid: percentage delta chemistry")

pct_pen_results <- summarise_penalized_results(
  lasso_obj = pct_lasso,
  ridge_obj = pct_ridge,
  label = "Percentage delta chemistry"
)

DT::datatable(
  pct_pen_results,
  options = list(pageLength = 30, scrollX = TRUE),
  caption = "Penalized viability model results: percentage delta chemistry"
)


### Comparsion Table ###

penalized_viability_results <- bind_rows(
  raw_pen_results,
  abs_pen_results,
  pct_pen_results
)

DT::datatable(
  penalized_viability_results,
  options = list(pageLength = 50, scrollX = TRUE),
  caption = "Combined LASSO and RIDGE-style viability model comparison"
)

penalized_viability_results %>%
  mutate(across(where(is.numeric), ~ round(.x, 3))) %>%
  knitr::kable(
    caption = "Combined LASSO and RIDGE-style viability model comparison"
  )
```


## Supp. Fig.: Soil Chemistry Ordination Plot
```{r, Soil Chemistry: Ordination Plot, cache=FALSE}
library(tidyverse)
library(vegan)
library(patchwork)

dir.create("outputs", showWarnings = FALSE)

# Variable Labels

chem_var_labs <- c(
  soil_moisture            = "Soil moisture",
  soil_water_hold_cap_log  = "Water holding capacity",
  soil_pH                  = "pH",
  soil_soluble_C_log       = "Total soluble C",
  soil_soluble_orgC_log    = "Soluble organic C",
  soil_soluble_inorgC_log  = "Soluble inorganic C",
  soil_N_log               = "Soluble N",
  soil_CN_ratio_log        = "C:N ratio",
  soil_Cl_log              = "Chloride",
  soil_K_log               = "Potassium",
  soil_P_log               = "Phosphorus"
)

# You can trim this list if the panel feels too busy.
key_chem_vars <- c(
  "soil_moisture",
  "soil_water_hold_cap_log",
  "soil_K_log",
  "soil_P_log",
  "soil_N_log",
  "soil_soluble_C_log",
  "soil_soluble_orgC_log",
  "soil_soluble_inorgC_log",
  "soil_CN_ratio_log",
  "soil_Cl_log",
  "soil_pH"
)

pretty_chem_var <- function(x) {
  x2 <- x %>%
    stringr::str_remove("^delta_") %>%
    stringr::str_remove("^pct_delta_") %>%
    stringr::str_remove("_pen_z$")

  dplyr::recode(x2, !!!chem_var_labs, .default = x2)
}


# PCoA ordination

pcoa_input <- soil_perm4_tsd %>%
  filter(!is.na(grp_tsd_dist)) %>%
  drop_na(all_of(chem_vars_z)) %>%
  mutate(
    years_since_death_f = factor(years_since_death_f, levels = c("2y", "3y")),
    distance_lab = factor(distance_lab, levels = c("0m", "15m")),
    grp_label = dplyr::recode(
      as.character(grp_tsd_dist),
      "2y_0m"  = "2y, 0 m",
      "2y_15m" = "2y, 15 m",
      "3y_0m"  = "3y, 0 m",
      "3y_15m" = "3y, 15 m"
    ),
    grp_label = factor(grp_label, levels = c("2y, 0 m", "2y, 15 m", "3y, 0 m", "3y, 15 m"))
  )

pcoa_dist <- vegan::vegdist(
  as.matrix(pcoa_input[, chem_vars_z]),
  method = "euclidean"
)

pcoa_fit <- cmdscale(
  pcoa_dist,
  k = 2,
  eig = TRUE
)

pcoa_var <- 100 * pcoa_fit$eig[1:2] / sum(pcoa_fit$eig[pcoa_fit$eig > 0])

pcoa_scores <- pcoa_input %>%
  bind_cols(
    tibble(
      PCoA1 = pcoa_fit$points[, 1],
      PCoA2 = pcoa_fit$points[, 2]
    )
  )

perm_r2 <- tryCatch(res_perm4_tsd$permanova$R2[1], error = function(e) NA_real_)
perm_p  <- tryCatch(res_perm4_tsd$permanova$`Pr(>F)`[1], error = function(e) NA_real_)

p_pcoa <- ggplot(
  pcoa_scores,
  aes(
    x = PCoA1,
    y = PCoA2,
    color = grp_label,
    shape = distance_lab
  )
) +
  stat_ellipse(
    aes(group = grp_label),
    type = "norm",
    level = 0.68,
    linewidth = 0.5,
    alpha = 0.55,
    show.legend = FALSE
  ) +
  geom_point(size = 2.7, alpha = 0.9) +
  labs(
    title = "Soil chemistry profiles",
    subtitle = paste0(
      "PERMANOVA R² = ", round(perm_r2, 3),
      "; p = ", format.pval(perm_p, digits = 2)
    ),
    x = paste0("PCoA1 (", round(pcoa_var[1], 1), "%)"),
    y = paste0("PCoA2 (", round(pcoa_var[2], 1), "%)"),
    color = "Group",
    shape = "Distance"
  ) +
  theme_bw(base_size = 11) +
  theme(
    panel.grid.minor = element_blank(),
    legend.position = "bottom",
    plot.title = element_text(face = "bold")
  )

supp_fig_soil_chem_pcoa <- p_pcoa +
  labs(
    title = NULL,
    subtitle = NULL
    ) +
  theme(
    plot.title = element_text(face = "bold"),
    plot.caption = element_text(hjust = 0, size = 9)
  )

supp_fig_soil_chem_pcoa

ggsave(
  filename = "PCoA.png",
  plot = supp_fig_soil_chem_pcoa,
  width = 7,
  height = 5.5,
  dpi = 600
)
```